# Proofs {#sec-appendix-a}

```{r}
#| label: setup-appa
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

@sec-ch02 states the rule this appendix implements: proofs run inline where they take less
than a page and move here otherwise, and nothing is asserted without either an argument or a
source locator. The two alternatives must not be blurred. Reproducing a useful special-case
derivation is not the same as reproducing the source theorem that licenses it.

The index therefore records a **claim-level proof status**. `proved-here` means the displayed
claim is established in this book; `partial-derivation` means this appendix establishes only
the named component and leaves the source theorem at its locator; `source-only` means the
published result is invoked under its stated conditions and no stand-alone proof is claimed
here. This appendix improves checkability, but it does not promise to replace every cited
proof.

## Index {#sec-appa-index}

```{r}
#| label: tbl-proof-index
#| tbl-cap: "Every numbered result and where its proof lives. Source: tables/T-proof-index.rds."
#| echo: false
knitr::kable(tbl("T-proof-index"))
```

Two things are worth reading off this table. Every result tagged **derived-here** has a
`proved-here` argument. Restated or adapted results can be `proved-here`,
`partial-derivation`, or `source-only`; none of those labels is an originality claim.

## A.1 The Dirichlet process {#sec-appa-dp}

@sec-ch15 defines the process and states three structural results from their sources. The
arguments are short enough to give in full, and giving them makes Part V self-contained.

### A.1.1 Stick-breaking {#sec-appa-stick}

**Source result** (@thm-stick, *restated from Sethuraman 1994*). Let $V_1, V_2, \dots$ be i.i.d.
$\mathrm{Beta}(1,\alpha)$, let $\phi_1, \phi_2, \dots$ be i.i.d. from the normalized base
measure $\bar G_0$ and independent of the $V$'s, and set

$$p_n = V_n \prod_{\ell < n} (1 - V_\ell), \qquad
P = \sum_{n=1}^{\infty} p_n \delta_{\phi_n} .$$

Then $P \sim \mathrm{DP}(\alpha \bar G_0)$.

**Proof status: partial derivation.** The two weight facts used later are proved below. The
finite-dimensional Dirichlet law of $P$ is Sethuraman's theorem and remains source-only.

*Proof of the two facts the chapter uses.* The weights sum to one almost surely. Writing
$R_N = \prod_{\ell \le N}(1 - V_\ell)$ for the unbroken remainder after $N$ breaks, induction
on the definition gives $\sum_{n \le N} p_n = 1 - R_N$. Since the $V_\ell$ are i.i.d. with
$\operatorname{E}[\log(1-V_\ell)] = \int_0^1 \log(1-v)\,\alpha(1-v)^{\alpha-1}\,dv
= -1/\alpha$, which is a strictly negative finite constant, $\log R_N$ is a random walk with negative drift and
$R_N \to 0$ almost surely. Hence $\sum_n p_n = 1$.

The mean weight sequence is geometric. $\operatorname{E}[V_n] = 1/(1+\alpha)$ and
$\operatorname{E}[1 - V_\ell] = \alpha/(1+\alpha)$, and independence across $\ell$ gives

$$\operatorname{E}[p_n] = \frac{1}{1+\alpha}\left(\frac{\alpha}{1+\alpha}\right)^{n-1}
= \frac{\alpha^{n-1}}{(1+\alpha)^n} .$$

So the expected weights decay geometrically at rate $\alpha/(1+\alpha)$, and the expected
mass left after $N$ atoms is $(\alpha/(1+\alpha))^N$, which is the truncation bound
@sec-ch15 uses. `V8` checks all three statements numerically (`THM-15-2`, `THM-15-2b`,
`THM-15-2c`).

That the resulting $P$ has Dirichlet finite-dimensional distributions is Sethuraman's
theorem and is not reproved here; the locator is in the index. □

### A.1.2 A draw is discrete with probability one {#sec-appa-discrete}

**Claim** (@prp-discrete, *restated; immediate from @thm-stick*).

*Proof.* By A.1.1 the random measure is $\sum_n p_n \delta_{\phi_n}$ with $\sum_n p_n = 1$
almost surely. A countable convex combination of point masses is a discrete measure. The
conclusion holds **however smooth $\bar G_0$ is**: continuity of the base measure controls
where the atoms fall, not whether there are atoms. That is the fact @sec-ch16 needs when it
warns that a DPM's mixing distribution is discrete even under a Gaussian base. □

### A.1.3 The Pólya urn and the CRP {#sec-appa-crp}

**Source result** (@thm-crp, *restated from Blackwell and MacQueen 1973*). If
$\theta_1, \theta_2, \dots$ follow the Pólya urn scheme

$$\theta_{n+1} \mid \theta_1, \dots, \theta_n \;\sim\;
\frac{\alpha}{\alpha + n}\,\bar G_0 + \frac{1}{\alpha + n}\sum_{i \le n} \delta_{\theta_i},$$

then the empirical measure converges almost surely to a discrete random measure $P$, and
$P \sim \mathrm{DP}(\alpha \bar G_0)$ with the $\theta$'s i.i.d. $P$ given $P$.

**Condition for the chapter's distinct-value reading.** The seating partition exists for
any base measure. Identifying “a new table” with “a new observed value” additionally requires
$\bar G_0$ to be nonatomic; otherwise a fresh base-measure draw can coincide with a value
already observed.

*Proof of the partition statement the book uses.* Read the scheme as a seating rule. Draw
$n+1$ joins an existing cluster of size $m_k$ with probability $m_k/(\alpha+n)$ and starts a
new one with probability $\alpha/(\alpha+n)$. The induced distribution on partitions is
therefore exchangeable and depends on the labels only through the cluster sizes — the
Chinese restaurant process. Marginalizing the urn over the $\phi$'s in A.1.1 gives the same
partition law, which is the correspondence @sec-ch15 uses to move between the two
representations. The almost-sure convergence and the identification of the limit as a DP are
Blackwell and MacQueen's martingale argument, whose locator is in the index. □

### A.1.4 Antoniak's law for the number of occupied clusters {#sec-appa-antoniak}

**Claim** (used in @sec-ch15 and @sec-ch16). Under the CRP with concentration $\alpha$, the
number $K_J$ of occupied tables among $J$ draws has

$$
\Pr(K_J=k\mid\alpha)=
\frac{|s(J,k)|\alpha^k}{(\alpha)_J},
\qquad (\alpha)_J=\alpha(\alpha+1)\cdots(\alpha+J-1),
$$

and

$$\operatorname{E}[K_J \mid \alpha] = \sum_{j=0}^{J-1} \frac{\alpha}{\alpha + j}
= \alpha\left[\psi(\alpha + J) - \psi(\alpha)\right],$$

where $\psi$ is the digamma function. If $\bar G_0$ is nonatomic, occupied tables and
distinct observed values coincide. Moreover, for fixed $J$, $K_J$ is sufficient for
$\alpha$ in the induced partition model.

*Proof.* Let $B_j$ indicate that draw $j$ starts a new table, $j = 1, \dots, J$, with
$B_1 = 1$. By the seating rule of A.1.3, $\Pr(B_{j+1} = 1 \mid \text{first } j) =
\alpha/(\alpha + j)$, which does **not** depend on the realized partition. So the $B$'s are
independent Bernoulli variables. Their probability-generating function is

$$
\operatorname{E}[t^{K_J}\mid\alpha]
=\prod_{j=0}^{J-1}\frac{\alpha t+j}{\alpha+j}
=\frac{\sum_{k=1}^{J}|s(J,k)|(\alpha t)^k}{(\alpha)_J},
$$

where the last equality is the rising-factorial expansion for the unsigned Stirling
numbers of the first kind. Extracting the coefficient of $t^k$ gives the pmf. Also
$K_J=\sum_j B_j$ gives the expectation, and $\psi(x+1)-\psi(x)=1/x$ gives the digamma form.

For sufficiency, the CRP probability of a particular partition with block sizes
$n_1,\ldots,n_k$ is

$$
\frac{\alpha^k}{(\alpha)_J}\prod_{h=1}^{k}(n_h-1)!.
$$

Its dependence on $\alpha$ is only through $k=K_J$, so the factorization criterion makes
$K_J$ sufficient for $\alpha$ in this partition model. □

The independence is the step worth noticing: it is what makes $\operatorname{E}[K_J]$
available in closed form, and it is why the elicitation of @sec-ch15 can invert the
relation to choose a prior on $\alpha$ from a belief about $K$. `V8` checks the identity and
the elicitation against it (`THM-15-4`, `THM-15-4b`, `THM-15-4c`).

## A.2 Identification of $G$ from a finite test {#sec-appa-partialid}

### A.2.1 What a finite test identifies {#sec-appa-i1}

**Claim** (@prp-partial-id, *restated from San Martín et al. 2011, Theorem 5*). Under the
source's anchor-item and availability conditions for the semiparametric Rasch model, the
observable law is generated by the item parameters and the following $I+1$ displayed basic
integral evaluations of $G$.

*Argument.* The observable is the distribution of the response vector
$\mathbf{U} \in \{0,1\}^I$, which is $2^I$ probabilities summing to one. Under local
independence and the Rasch form, the probability of any pattern with total score $r$ is

$$\Pr(\mathbf{U} = \mathbf{u}) = \int \prod_{i} \frac{e^{u_i(\theta - \beta_i)}}
{1 + e^{\theta - \beta_i}} \, dG(\theta),$$

Writing $r=\sum_i u_i$, the pattern probability equals
$\exp(-\sum_i u_i\beta_i)\mu_r$, so the whole observable law is a function of the item
parameters and of the $I+1$ evaluations

$$\mu_r = \int \frac{e^{r\theta}}{\prod_i (1 + e^{\theta - \beta_i})}\, dG(\theta),
\qquad r = 0, 1, \dots, I,$$

one for each attainable total score. These evaluations are not unconstrained independent
coordinates. Summing all pattern probabilities gives the identity

$$
\sum_{r=0}^{I}e_r(e^{-\beta_1},\ldots,e^{-\beta_I})\mu_r=1,
$$

where $e_r$ is the $r$th elementary symmetric polynomial. Two latent distributions
producing the same displayed evaluations are observationally equivalent. San Martín et al.
invert the observable map under their source conditions; this argument establishes the
factorization and identity, not an independent-dimension or universal “no more” claim.

The full inversion under the anchor restriction is their Theorem 5 and remains source-only.
Their Theorem 3
concerns the Rasch **Poisson counts** model and gives full identification of $\beta$ and $G$
with at least two probes; conflating the two is the manuscript error recorded as `C-014`. □

### A.2.2 A centred base measure does not centre the realized DPM {#sec-appa-dpmconstraint}

**Claim** (@prp-dpm-constraint, *adapted from San Martín et al. 2011*). If $G \sim
\mathrm{DP}(\alpha G_0)$, $\int \theta \, dG_0(\theta) = 0$, and
$0<\operatorname{Var}_{G_0}(\theta)<\infty$, then the realized mean
$M_G=\int\theta\,dG(\theta)$ is nondegenerate, with

$$
\operatorname{E}(M_G)=0,
\qquad
\operatorname{Var}(M_G)=\frac{\operatorname{Var}_{G_0}(\theta)}{1+\alpha}.
$$

*Proof.* By A.1.1, $\int \theta\,dG(\theta) = \sum_n p_n \phi_n$ with the $\phi_n$ i.i.d.
from $\bar G_0$ and the $p_n$ the stick-breaking weights, independent of them. Conditional on
the weights, this is a weighted sum of mean-zero draws, so its conditional mean is zero and
its conditional variance is $\sigma^2_{G_0} \sum_n p_n^2$, which is finite and strictly
positive under the stated assumption because the weights sum to one. The realized mean is therefore a non-degenerate
mean-zero random variable, not the constant zero.

Two consequences the book uses. Centring the *base measure* is not the same restriction as
centring the *realized* distribution, so a location constraint imposed on the base does not
identify the model; and $\operatorname{E}[\sum_n p_n^2] = 1/(1+\alpha)$ gives the displayed
unconditional variance. Thus the discrepancy shrinks as $\alpha$ grows and is largest
exactly where the prior is most flexible. □

## A.3 Reliability, separation, and the design constant {#sec-appa-reliability}

### A.3.1 Separation and strata {#sec-appa-separation}

**Claim** (@prp-separation, *restated from Wright and Masters 1982, § 5.5*). With
$\bar w = \sigma^2/(\sigma^2 + \mathrm{MSEM})$, the person separation index and the number of
strata are

$$S = \frac{\sigma}{\mathrm{RMSEM}} = \sqrt{\frac{\bar w}{1 - \bar w}},
\qquad H = \frac{4S + 1}{3} .$$

*Proof of the separation identity.* $\bar w/(1-\bar w) = \sigma^2/\mathrm{MSEM}$, and
taking square roots gives $S = \sigma/\mathrm{RMSEM}$.

**The strata formula is not derived here, and Wright and Masters do not derive it either.**
They state $H = (4G+1)/3$ for both the item and the person index and give the convention it
encodes — statistically distinct strata have centres **three calibration or measurement
errors apart** — without showing where the constants come from. That convention is the
substantive content, it is what the manuscript had used without a source, and it is recorded
as `C-003`. This book restates the formula on their authority; a reader wanting the constants
justified will not find it in the source, and inventing a derivation here would misrepresent
what is known.

This is the identity that exposes `C-001`. Computing $S$ from a *rounded* RMSEM rather than
from the exact target reliability is what produces the manuscript's $1.3$, $1.4$ and $3.3$
where the exact column is $1.0$, $1.2$, $1.5$, $2.0$, $3.0$. `V8` checks both readings
(`PROP-08-2`, `C-001`, `C-001b`). □

### A.3.2 The design constant, and the condition it needs {#sec-appa-cstar}

**Claim** (@prp-cstar, deferred from @sec-ch09). Fix the item count, difficulties, baseline
discriminations, and population distribution. Let $c>0$ multiply every discrimination, and
suppose both reliability functions have unique strictly increasing branches through the
target. The discrimination multiplier required under the MSEM definition is at least the
multiplier required under the average-information definition.

*Proof.* At a fixed value of the discrimination multiplier $c$, let
$\mathcal{J}_c(\theta)$ be the information computed with both the $c$-scaled slopes and the
resulting $c$-dependent response probabilities. MSEM is the mean of the
conditional error variances, $\operatorname{E}_\theta[1/\mathcal{J}_c(\theta)]$; the
average-information summary is the reciprocal of the mean information,
$1/\operatorname{E}_\theta[\mathcal{J}_c(\theta)]$. Since $x \mapsto 1/x$ is strictly convex on
$(0,\infty)$, Jensen gives

$$\operatorname{E}_\theta\!\left[\frac{1}{\mathcal{J}_c(\theta)}\right] \;\ge\;
\frac{1}{\operatorname{E}_\theta[\mathcal{J}_c(\theta)]},$$

with equality if and only if $\mathcal{J}_c(\theta)$ is constant in $\theta$. So the MSEM
error is the larger and the MSEM reliability is no larger at every common $c$. On the
assumed unique increasing branches, the inverse image of a common target is therefore
reached no earlier by $\bar w(c)$ than by $\tilde\rho(c)$, so
$c^\star_{\bar w}\ge c^\star_{\tilde\rho}$. The strictness of the inequality away from constant information is why
the two definitions diverge exactly in the item-response setting, where information depends
on $\theta$ by construction (@sec-ch07).

The uniqueness condition is not cosmetic: without a single strictly increasing branch the
inverse image is not a point and “the required discrimination multiplier” is not well defined. `V8`
checks the inequality and, separately, that scaled 2PL information is *not* obtained by
freezing the response probabilities (`PROP-09-1`, `PROP-09-1b`). □

## A.4 Posterior summaries {#sec-appa-summaries}

### A.4.1 Rank of the mean against mean of the rank {#sec-appa-rankofmean}

**Claim** (@prp-rank-of-mean, *restated from Shen and Louis 1998*). The rank of the posterior
mean and the posterior mean of the rank differ in **value** in general, and differ in
**ordering** only outside stochastically ordered regimes.

*Proof.* For the value statement, the expected rank of unit $k$ is
$\bar R_k = \sum_q \Pr(\theta_q \le \theta_k \mid \mathbf{u})$, which for independent
posteriors is a sum of $K$ probabilities and takes values in $[1, K]$ on a continuous scale,
while $\operatorname{rank}(\eta_k)$ is an integer in $\{1, \dots, K\}$. They are different
objects and coincide only by accident.

For the ordering statement, the tempting term-by-term comparison of the self and cross
summands is false. The valid result is Shen and Louis's Theorem 2: under their stochastic
ordering condition, posterior means, expected ranks, and the relevant rank summaries induce
the same order. That published theorem is invoked at its locator rather than replaced by the
invalid summand argument. The converse fails: two normals with different variances are
**not** stochastically ordered, their standardized arguments cross, and the expected-rank
ordering can then disagree with the ordering of the means.

@sec-ch12 supplies the case that matters for this book. In a common-form Rasch model the
posterior depends on the data only through the total score and the family is ordered in it,
so the regime is the stochastically ordered one and the orderings agree — which is why the
correction recorded there withdrew an earlier claim that differential shrinkage reorders. □

### A.4.2 The triple-goal construction {#sec-appa-gr}

**Claim** (@thm-gr, *restated from Shen and Louis 1998, §§ 2.1, 2.3, 2.4*). The triple-goal
estimator is built in three steps: form $\bar G_K(t) = K^{-1}\sum_k
\Pr(\theta_k \le t \mid \mathbf{u})$ and take its $K$ midpoint quantiles $\hat U_j =
\bar G_K^{-1}\!\left(\frac{2j-1}{2K}\right)$; compute expected ranks $\bar R_k$ and
discretize them to integer ranks $\hat R_k$; and set $\hat\theta_k = \hat U_{\hat R_k}$.

*What each step optimizes.* Step 1: the unrestricted pointwise squared-error Bayes action for
the ensemble EDF is the continuous posterior-mean function $\bar G_K$. When the action is
restricted to empirical CDFs with $K$ equal masses, the ISEL-optimal discrete action is
$\hat G_K$, whose atoms are the midpoint quantiles $\hat U_j$ of $\bar G_K$. In general
$\hat G_K\ne\bar G_K$; discretization approximates the continuous Bayes action and does not
carry it exactly. Step 2: the expected rank minimizes squared-error rank loss, by the same
posterior-mean argument applied to $R_k$; the source construction converts those real-valued
actions to a permutation. Step 3 assigns the ordered atoms by that expected-rank permutation.
Separately, the rearrangement inequality says the coordinate-SEL-minimizing permutation pairs
the ordered atoms with the ordered posterior means. These are the same assignment only when
the expected-rank order and posterior-mean order agree.

Two things follow, both used in @sec-ch19. Every permitted step-3 assignment is a permutation
of the atoms and therefore carries the **discrete action $\hat G_K$** exactly. It does not
carry the continuous $\bar G_K$ exactly. And $\bar G_K(t)$ is a *probability* — an average
of $K$ posterior probabilities — not a trait estimate. □

## A.5 What is deliberately not here {#sec-appa-notes}

The proofs of Sethuraman's finite-dimensional DP theorem, Blackwell and MacQueen's martingale
convergence, San Martín et al.'s inversion theorem, Ghosh's general constrained-Bayes theorem,
and Shen and Louis's stochastic-order theorem are not reproduced. Each remains `source-only`
at the locator in the claim-level inventory. What is established above is narrower: the
weight identities, partition/Antoniak algebra, finite-test factorization and identity,
realized-DP mean variance, reliability ordering at common discrimination scale, and the
correct distinction among $\bar G_K$, $\hat G_K$, rank loss, and conditional coordinate-SEL
assignment.
