# Notation Crosswalk {#sec-appendix-b}

```{r}
#| label: setup-appb
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

A reader of this book will move between it and half a dozen other treatments, and the
symbols do not agree. This appendix exists so that the move does not silently change what a
formula says.

Every cell below was read in the source at the locator recorded in
`manifest/part-vii-source-receipts.csv` — the model-definition passage where the symbol is
introduced — rather than recalled. A dash means the source has no symbol for that concept in
the passage read, not that one was overlooked.

## The crosswalk {#sec-appb-table}

```{r}
#| label: tbl-crosswalk
#| tbl-cap: "This book's symbols against six treatments, read at each source's model definition. Source: tables/T-crosswalk.rds."
#| echo: false
knitr::kable(tbl("T-crosswalk"))
```

The bold entries are the ones that will cost a careless reader an hour. They are set out
below.

## Five collisions {#sec-appb-collisions}

```{r}
#| label: tbl-collisions
#| tbl-cap: "Symbols that mean different things in different sources. Source: tables/T-collisions.rds."
#| echo: false
knitr::kable(tbl("T-collisions"))
```

### $I$ is a count here and a reliability in Lee et al. {#sec-appb-I}

This is the one to watch, because @sec-ch21 puts both papers on the same page. In this book
$I$ is the number of items, an integer that indexes the identification result — under its
anchored Rasch conditions, the observable law is generated by $I+1$ displayed integral
evaluations of $G$, subject to their identities. In Lee et al. [-@lee_improving_2025]
$I$ is *informativeness*,

$$I = \frac{\sigma^2}{\sigma^2 + \exp\!\left(\tfrac{1}{J}\sum_j \ln \widehat{se}^2_j\right)}
\;\in\; [0,1],$$

an average reliability of the site-level maximum likelihood estimates, and the single most
influential factor in their study. A sentence like "performance improves with $I$" is true in
both papers and means two different things.

### $\eta$ is an ability in Paganin et al. and a posterior mean here {#sec-appb-eta}

Paganin et al. [-@paganin_computational_2022] write the 2PL as
$\operatorname{logit}(\pi_{ij}) = \lambda_i(\eta_j - \beta_i)$, so their $\eta_j$ is this
book's $\theta_p$. This book uses $\eta_p$ for the **posterior mean** of $\theta_p$, on every
page of @sec-ch11 and Part VI. The two are related by an integral, not by a change of letter.

The collision is worth flagging precisely because the rest of their notation agrees with this
book's: $\lambda_i$ for discrimination, $\beta_i$ for difficulty, $G$ for the latent
distribution, $\alpha$ for the DP concentration. That agreement is not accidental — the
companion package's parameterization draws on theirs — and it makes the one disagreement
easier to miss.

Their slope-intercept form $\lambda_i\eta_j + \gamma_i$ with $\gamma_i = -\lambda_i\beta_i$
is the same reparameterization Baker and Kim write as $\zeta_i = -\alpha_i\beta_i$.

### $\alpha$ is a discrimination in Baker and Kim {#sec-appb-alpha}

Baker and Kim [-@baker_item_2004] write the two-parameter normal ogive as
$Z_{ij} = \alpha_i(\theta_j - \beta_i)$, so $\alpha_i$ is the item discrimination. In this
book and in the Bayesian nonparametric literature $\alpha$ is the Dirichlet process
concentration parameter, and it never carries an item subscript. @sec-ch15 and @sec-ch16 use
it only in that sense; discrimination is always $\lambda_i$.

Baker and Kim also warn, at the same locator, that $Z_{ij} = \alpha_i\theta_j - \gamma_i$ is
an "atypical parameterization of item difficulty" that appears in the Gibbs-sampling
literature. A reader importing code from that literature should check which $\gamma$ is meant.

### Fox reverses both indices {#sec-appb-fox}

Fox [-@fox_bayesian_2010] indexes **persons** by $i$ and **items** by $k$: the response is
$Y_{ik}$, the ability is $\theta_i$, and the item parameters are collected in
$\boldsymbol\xi$ with the population hyperparameters in $\boldsymbol\theta_P$. This book
indexes persons by $p$ and items by $i$.

Every double-indexed formula therefore transposes. The shrinkage discussion in his § 2.1 —
that the posterior mean is a combination of the prior mean and a likelihood-based estimate,
shrunk more when the population parameters are informative about $\theta_i$ — is the same
statement as @sec-ch11's, with $i$ where this book writes $p$.

### $\lambda_p$: the one departure this book makes {#sec-appb-lambda}

Shen and Louis [-@shen_triple-goal_1998] write the posterior variance of unit $k$ as
$\lambda_k$. Under decision `TD-3` this book carries discrimination as $\lambda_i$ on most
pages of Parts III and VI, so the two would collide constantly.

**Resolved (`DECISIONS` a6): $\lambda_i$ keeps the simulation book's meaning and the
posterior variance becomes $v_p$.** This is the only place where this book departs from a
source's notation rather than merely differing from it, and it is stated at first use in
@sec-ch11, again at @sec-ch17 and @sec-ch18, and here.

## Reading a formula across the boundary {#sec-appb-practice}

Three habits cover most of the risk.

**Check the index order before trusting a double subscript.** Of the five sources with a
response variable, this book and Debelak et al. write person-then-item, Fox writes
person-then-item with the *letters* reversed, and Paganin et al. write item-then-person
($y_{ij}$, $i$ the item). The symbol $y_{21}$ means different cells in different papers.

**Check whether a Greek letter is an item parameter or a hyperparameter.** $\alpha$, $\gamma$
and $\lambda$ each appear on both sides of that line somewhere in this literature.

**Check whether a latent distribution is called a prior or an assumption.** Lee et al. attach
a footnote to exactly this point: they call $G$ a prior distribution to match Bayesian
hierarchical practice, and note that in a frequentist frame the same object is a
distributional assumption about $\tau_j$. This book uses "latent trait distribution" for the
object and reserves "prior" for the Bayesian reading, but the two literatures it draws on do
not agree, and the disagreement is terminological rather than substantive.

## The manuscript {#sec-appb-manuscript}

The submitted manuscript's notation is this book's, with two exceptions already recorded as
corrections rather than as translation problems: its Appendix D describes an item prior the
fitted model does not use (`C-013`), and its Appendix F attributes the partial-identification
result to the wrong theorem (`C-014`). Neither is a notation difference, so neither appears in
the crosswalk. A reader moving between the manuscript and this book needs @sec-appendix-c, not
this appendix.
