# Corrections to the Submitted Manuscript {#sec-appendix-c}

```{r}
#| label: setup-appc
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
cx  <- tbl("T-corrections")
cxi <- tbl("T-corrections-index")
```

This appendix is the revision's worklist. It collects every place where writing this book
required going back to a primary source and finding that the submitted manuscript —
APM-25-04-056, *Targeting Toward Inferential Goals in Bayesian Rasch Models for Estimating
Person-Specific Latent Traits* — says something wrong, imprecise, or merely unclear.

It exists because the book was written under a rule stated in the front matter: a claim enters
only when the source that established it has been read at the locator cited, not when the
manuscript's or a reviewer's summary of it has been read. Applying that rule produced
findings. Rather than fix them silently in the book and leave the paper unchanged, each
one is recorded here with the source consulted, so the revision can be worked through
entry by entry and the response letter can point at specific text.

## What counts as an entry, and what the severities mean

An entry is **not** a difference of emphasis or a place where the manuscript is terser
than the book. It is a statement whose truth value or attribution changes once the source
is opened.

| Severity | Meaning | Count |
|---|---|---|
| **moderate** | A substantive statement is wrong, or a citation attributes a result to a source that does not contain it. Fixing it changes what the paper claims | `r sum(cx$Severity == "moderate")` |
| **minor** | A number, a formula, or a piece of wording is wrong or imprecise, but the surrounding argument survives the fix | `r sum(cx$Severity == "minor")` |
| **none** | Checked against the primary source and found **sound**. Recorded so the revision does not spend effort re-verifying, and so a reviewer can see the check was made | `r sum(cx$Severity == "none")` |

The `none` rows are not padding. Three of them are places where an earlier draft of this
appendix asserted an error that a second reading withdrew; keeping them makes the audit
trail honest and stops the same false alarm from being raised twice. `C-007` is the
clearest case — a first pass mixed rows and units and produced a correction that was
itself wrong.

`r sum(cx$ID %in% c("C-010", "C-002", "C-003"))` entries are limitations rather than
errors: `C-010` records that the semiparametric
**2PL** identification problem is open, which the manuscript does not disclose, and
`C-002` and `C-003` record provenance that was previously unchecked and is now sourced.

## Index

```{r}
#| label: tbl-corrections-index
#| tbl-cap: "Every correction, ordered by severity then chapter. Source: tables/T-corrections-index.rds."
#| echo: false
knitr::kable(cxi)
```

Where an entry required a decision that only the author could make — because more than one
fix was defensible — the ruling and its date are recorded with the entry.
`r sum(cxi$Ruled != "—")` entries carry rulings; the rest have a single defensible fix and
needed none.

## The entries

Each entry gives, in order: where in the manuscript, what it says, why that is wrong (or
why it is sound), the source consulted at its locator, what the revision should do, and
the author's ruling.

```{r}
#| label: appc-entries
#| echo: false
#| results: asis
for (i in seq_len(nrow(cx))) {
  r <- cx[i, ]
  cat(sprintf("\n### %s — %s {#sec-appc-%s}\n\n",
              r$ID, r$`Where in the manuscript`, tolower(r$ID)))
  cat(sprintf("*Severity: **%s**. Treated in %s of this book.*\n\n",
              r$Severity, paste0("ch. ", r$`Book chapter`)))
  cat("> ", gsub("\n", " ", r$`What it says`), "\n\n", sep = "")
  cat("**Why that is wrong.** ", gsub("\n", " ", r$`Why that is wrong`), "\n\n", sep = "")
  cat("**Source consulted.** ", gsub("\n", " ", r$`Source consulted`), "\n\n", sep = "")
  cat("**What the revision should do.** ", gsub("\n", " ", r$`What the revision should do`),
      "\n\n", sep = "")
  cat("**Author ruling.** ", gsub("\n", " ", r$`Author ruling`), "\n\n", sep = "")
}
```

## What is not here {#sec-appc-not-here}

Three classes of thing are deliberately absent.

**Results.** No entry concerns whether a simulated condition came out one way or another.
That boundary is the companion simulation study's, and this book does not cross it; the
front matter sets out which document is authority for what.

**Presentation.** Places where the manuscript would read better are not corrections. The
three that came out of Part VI and were ratified for the revision are recorded in
`DECISIONS.md` rather than here: naming $G_N$ as the realized empirical distribution
(@sec-ch17), stating that $\bar G_N$ is a probability rather than a trait estimate
(@sec-ch19), and saying in one sentence that test length was the instrument by which
reliability was set rather than a crossed design factor (@sec-ch09).

**Errors in this book.** They exist, and external review found them. They are recorded
in the project's step logs, which are not part of the published release, rather than
here, because this appendix is about the manuscript. The distinction matters for one reason worth stating plainly: several of the
corrections above were found only because a check was written that *could* fail, and the
same discipline applied to the book's own prose refuted claims this book had made. An
appendix that listed only the manuscript's errors would misrepresent where the errors were.

## Using this appendix for the response letter

The entries map onto the reviewers' comments unevenly. `C-009` is the substance of the
answer to Reviewer 2's third comment and should be quoted in the response, not merely
fixed in the text: the identification warrant the manuscript implicitly claims is narrower
than stated, and saying so is stronger than defending the wider claim. `C-014` and `C-010`
both bear on Reviewer 2's identifiability question and are best answered together, since
one is a misattributed theorem and the other an undisclosed open problem in the same
literature. `C-008` is a reference-list repair that Reviewer 2's overlap comment makes
conspicuous. The remaining entries are text repairs that need no comment in the letter.
