# Annotated Bibliography {#sec-appendix-d}

```{r}
#| label: setup-appd
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
ab <- tbl("T-annotated-bib")
tr <- tbl("T-bib-tiers")
n_independent <- tr$Entries[tr$Tier == "primary-read"]
n_programme <- tr$Entries[tr$Tier == "programme-artifact-read"]
n_read <- n_independent + n_programme
n_all  <- sum(tr$Entries)
unbound <- ab[ab$Evidence == "none", , drop = FALSE]
n_unbound <- nrow(unbound)
unbound_names <- if (n_unbound) paste(unbound$Source, collapse = ", ") else "none"
cv <- read.csv(file.path("..", "..", "manifest", "citation-verification.csv"),
               stringsAsFactors = FALSE, check.names = FALSE)
n_cv <- nrow(cv)
n_loc <- sum(cv$level == "locator")
n_corr <- sum(grepl("CORRECTION", cv$verdict))
```

An annotated bibliography is a document made **entirely** of claims about sources. That is
the exact class of claim four external review cycles found errors in — a theorem attributed
to the wrong number, an absence certified in a paper whose discussion states the opposite, a
condition dropped from a proposition. So this appendix is scoped by a number rather than by a
plan.

**`r n_read` of `r n_all` bibliography entries were opened at a locator and are annotated:**
`r n_independent` `primary-read` works from the independent literature and `r n_programme`
`programme-artifact-read` works produced by this research programme. The second count is
deliberately separate. Reading this programme's own book, manuscript, package, code, or frozen
table can establish what that artifact says or does; it cannot supply independent corroboration.
The rest are held, or cited as secondary treatments, and the register says which.

Nothing below is written from memory. Each annotation is assembled by
`code/R/17-annotated-bibliography.R` from what the project has **already verified** about
that source — a semantic receipt, a numbered result sourced to it, a correction it grounds,
a citation verification recorded at a locator, a bounded narrative receipt, or a programme
artifact receipt — and the `Evidence` column names which. The `Tier` and `Authority` columns
prevent a programme artifact from being displayed as independent literature.

Citation verification is the narrowest evidence route for an external source.
`manifest/citation-verification.csv` records what happened when a citation was checked
against the source itself: which sites were involved, which file was opened, what was found
there, and on what date. Only rows marked `level = locator` **and** confirmed at that
locator are admitted as evidence here. Rows marked `ATTRIBUTION SOUND` compared a title and
its distinctive terms against the citing claim without opening a passage; they are refused,
and the sources they cover stay `primary-held`, so they never reach this appendix at all.

That register also carries the checks that did **not** come out clean. Of `r n_cv` checks,
`r n_loc` were made at a locator and **`r n_corr` of those forced a correction to this
book** — a rate worth stating, because it is the argument for doing the checks at all. They
were not all the same kind of error:

- **A locator that points at the wrong page.** van der Ark (2025) was cited in @sec-ch08 at
  two page numbers and both were one page late — the Lord and Novick lower-bound result is
  on p. 1679, not p. 1680, and the $\lambda_3 = \alpha$ identity is in § 3.7 on p. 1686, not
  p. 1687, where an unrelated simulation section begins. The content was exactly as
  described; only the locators were wrong.
- **A number that was not the source's number.** @sec-ch01 gave a Grade 9 battery as
  reliable at ".83 to .92 overall"; Abedi's Table 9 runs .80 to .92, a bound that excluded
  four of seven subscales. The corrected figures make the point better, since the table also
  gives .53 to .83 for English language learners.
- **A claim stronger than the source's.** @sec-ch08 put Hussey et al.'s excess of $\alpha$
  values "just above" $.70$; they locate it *at* $.70$, which is also their title.
- **A citation the source does not support at all.** @sec-ch08 grouped Cronbach and
  Shavelson (2004) with Sijtsma as making the point about Nunnally's threshold. A full-text
  scan finds no occurrence of *Nunnally*, *.70*, or *threshold* anywhere in it.
- **An attribution to the wrong work by the right author.** @sec-ch12 credited the held
  Robbins reprint with the compound-decision framing; that phrase appears in it only in its
  reference list, pointing at a different paper.
- **A disclaimer that was itself wrong.** @sec-ch08 declined to give a page for
  @eq-alpha because "the held copy is unpaginated OCR". The held copy is a clean paginated
  scan and the equation is Cronbach's (2) on p. 299.

Only the first of those six is the error a careless reader would predict. The rest are
failures of *what was claimed*, not of where it was looked for, and none would have been
caught by any check short of opening the source.

**One thing is deliberately *not* an evidence type.** `manifest/citation-context-match.csv`
records, for 66 narrative citations, which page of a PDF shares the most words with the
citing sentence. Review 021 used that table to promote 60 sources to `primary-read`; the
promotion was reverted, because keyword overlap is not reading — several matches landed on
OCR noise, and one covered a source this book cites at equation level. The table survives
as a navigation aid for whoever reads those sources next, it declares `is_reading = NO` on
every row, and `V1` refuses to let it claim otherwise.

## What the tiers mean {#sec-appd-tiers}

```{r}
#| label: tbl-bib-tiers
#| tbl-cap: "The bibliography by verification tier. Source: tables/T-bib-tiers.rds."
#| echo: false
knitr::kable(tr)
```

`primary-held` is not a criticism of a source; it is a statement about this book. It means the
work is in one of the nine reference libraries and was available, but no claim here rests on
having read it at a locator. Where such a work is named in the text, the convention
(@sec-ch21, @sec-ch22) is to name it without an `@` citation and say that what was read is
another author's use of it.

`programme-artifact-read` is a different statement: the artifact was opened directly, but it
belongs to the same research programme. Its receipt must state both what was checked and the
authority boundary. It may establish cross-book or code-to-prose fidelity; it may not be counted
as independent confirmation.

The `r n_read` annotated below are the sources and programme artifacts this book reads directly.

## The annotations {#sec-appd-entries}

Ordered by the literature group each belongs to, which is assigned by the first chapter that
cites it rather than by judgement. `What was verified` is the receipt's own text where one
exists, so it records what was checked, not what the source is generally about.

```{r}
#| label: appd-entries
#| echo: false
#| results: asis
for (g in unique(ab$Group)) {
  cat(sprintf("\n### %s {#sec-appd-%s}\n\n", g,
              tolower(gsub("[^a-z]+", "-", tolower(g)))))
  sub <- ab[ab$Group == g, , drop = FALSE]
  for (i in seq_len(nrow(sub))) {
    r <- sub[i, ]
    cat(sprintf("**%s.** *%s*\n\n", r$Source, r$Title))
    cat(sprintf("> %s\n\n", gsub("\n", " ", r$`What was verified`)))
    cat(sprintf("*%s · %s · cited in ch. %s · read at %s · %s · read on %s*\n\n",
                r$Tier, r$Authority, r$`Cited in`, r$At, r$Evidence, r$`Read on`))
  }
}
```

## What this appendix exposes {#sec-appd-gaps}

Generating rather than writing it surfaced two things worth stating plainly.

**`r n_unbound` locator-read annotations are unbound (`r unbound_names`).** This is an
executable release condition, not an aspiration: the generator stops if a locator-read entry
lacks a receipt. The inherited six were resolved without inventing evidence. Lo (1984) now has
a bounded Chapter 15 narrative receipt. Five programme-owned works that already had direct
artifact reads now carry programme receipts, and all eight programme-owned reads are classified
under the separate `programme-artifact-read` tier.

Conoyer's correction binding remains explicit as well: corrections join through the
bibliography's `verified_claims`, rather than by searching a prose locator for a BibTeX key.

**The annotations are uneven in length, and the unevenness is informative.** Sources with long
entries are those whose receipts record several distinct verified claims — Lee et al. (2025),
Masters (1982), Gu and Xu (2020). Sources with a one-line entry supply a single result. The
length of an annotation therefore measures how much of this book rests on that source, which
is more useful than a uniform paragraph would be.

## What is deliberately not here {#sec-appd-limits}

**No evaluations.** This appendix does not say whether a source is good, influential, or
superseded. It says what was read in it and where. A judgement about a literature would need a
survey this book has not conducted, and @sec-ch23 records where that absence matters.

**No entries for the `r tr$Entries[tr$Tier == "primary-held"]` held-but-unread works.** Listing
them by title would fill pages and make none of them checkable. They are in
`refs/references.bib` with their tier, and `code/R/03-citation-check.R` refuses any numbered
result that cites one as though it had been read. The same checker validates every programme
receipt, narrative receipt, date, bibliography key, and `verified_claims` binding.

**No reading plan.** Which of the held works should be read next is a question about the next
piece of research, not about this book. The live, owner-assigned acquisition debt is shown
separately in @sec-appendix-e; an entry there is a request to acquire or disambiguate a work,
not evidence that it has been read.
