# Reference Acquisition Ledger {#sec-appendix-e}

```{r}
#| label: setup-appe
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
ledger <- tbl("T-acquisition-ledger")
debt <- tbl("T-acquisition-debt")
stopifnot(nrow(ledger) == 33L, !anyDuplicated(ledger$Key))
n_outstanding <- sum(ledger$Status == "outstanding")
n_identity <- sum(ledger$Status == "confirm identity first")
```

This appendix is the rendered view of the live acquisition source
`refs/acquisition-ledger.csv`. It contains **`r nrow(ledger)` entries**: `r n_outstanding`
ordinary outstanding acquisitions and `r n_identity` entries whose bibliographic identity
must be confirmed before acquisition. The source CSV, not this prose, is authoritative.

The ledger is current work debt, not a historical claim that the project still lacks every
source named in early planning documents. A work can remain here because the preferred
edition or primary route is still wanted even when a secondary route is already sufficient
for the bounded statement in the book. Conversely, an `outstanding` row does not authorize
the book to cite the work as read; V1 and the bibliography tiers govern that separately.

## Status debt {#sec-appe-status}

```{r}
#| label: tbl-acquisition-debt
#| tbl-cap: "Open acquisition entries by current status and criticality. Source: tables/T-acquisition-debt.rds."
#| echo: false
knitr::kable(debt)
```

The three `uncertain` criticality rows are the same three `confirm identity first` rows.
They are not acquisition requests until author, title, and year are disambiguated. A named
R5 route records the secondary source that can support a narrower statement; it does not
turn the absent primary into a `primary-read` source.

## Per-entry ledger {#sec-appe-entries}

```{r}
#| label: tbl-acquisition-ledger
#| tbl-cap: "The complete 33-entry acquisition ledger. Locator is the stable source-CSV line for each entry. Source: tables/T-acquisition-ledger.rds."
#| echo: false
knitr::kable(ledger)
```

Every row has a key, target chapter, status, owner, and a locator back to the source CSV.
Blank route or date fields mean “not yet recorded”; they are displayed as blanks rather
than the literal token `NA`. Closing a row requires updating the source CSV and, if the work
will support a claim, separately adding the bibliography tier, locator, and receipt demanded
by V1. The next build then regenerates this appendix and reconciles the row count exactly.

## Retirement rule {#sec-appe-retirement}

An entry leaves this ledger only when one of three things happens: the intended source is
acquired and processed into the reference inventory; a verified substitute makes the
preferred source unnecessary and the decision is recorded; or the intended identity is
shown not to exist. Merely finishing a chapter does not close its acquisition debt.
