# Build and Verification Report {#sec-appendix-g}

```{r}
#| label: setup-appg
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
bld <- tbl("T-build-report")
ver <- tbl("T-verifiers")
bootstrap <- identical(Sys.getenv("DPMIRT_APPENDIX_G_BOOTSTRAP"), "1")
val <- function(q) {
  z <- bld$Value[bld$Quantity == q]
  if (!length(z) && bootstrap) "pending" else z
}
if (bootstrap && !"Current output" %in% names(ver) &&
    "Output file" %in% names(ver))
  names(ver)[names(ver) == "Output file"] <- "Current output"
if (bootstrap && !"Run token" %in% names(ver)) ver$`Run token` <- "bootstrap-pending"
if (!"Current output" %in% names(ver)) stop("T-verifiers lacks current output names")
if (!"Run token" %in% names(ver)) stop("T-verifiers lacks the current-run token")
run_tokens <- unique(ver$`Run token`)
if (length(run_tokens) != 1L || !nzchar(run_tokens))
  stop("T-verifiers does not bind all V1--V9 rows to one run token")
run_token <- run_tokens[[1]]
ver_view <- ver[, c("Verifier", "Current output", "Issues", "Advisories"), drop = FALSE]
## The claim-level table is written by the same generator as the two above; on the
## bootstrap pass it does not exist yet, and the report pass is what readers see.
claim_checks <- tryCatch(tbl("T-claim-checks"), error = function(e) NULL)
if (is.null(claim_checks)) {
  if (!bootstrap) stop("T-claim-checks is required for the report pass")
  claim_checks <- data.frame(Checker = "pending", `What it audits` = "pending",
                             `Current output` = "pending", Checks = NA_integer_,
                             `Negative fixtures` = NA_integer_, Issues = NA_integer_,
                             check.names = FALSE)
}
```

## The current release receipt {#sec-appg-summary}

**This report is current-run evidence, not an inventory of whatever happened to be
left in `verification/`.** The release driver bound
`r val("Current-run verifier output receipts")` required V1--V9 outputs to build token
`r run_token`; every receipt matched that token, byte count, and SHA-256 before these
tables were frozen. The current issue total is `r sum(ver$Issues)`; V9 separately has
`r sum(ver$Advisories)` lexical coverage advisories, which are intentionally
non-blocking.

`code/R/16b-build-report.R` refuses to generate this appendix if a required output is
missing, unreadable, named differently from the contract, receipted under another run,
or changed after its receipt was written. Three safe mutation fixtures establish the
negative cases: a zero-row issue file with an old token, a receipted zero-row file
changed afterward, and a missing required output. A zero therefore means “the named
current-run check passed,” not “a file was absent” and not “an older empty CSV survived.”

## What the book contains {#sec-appg-contents}

```{r}
#| label: tbl-build-report
#| tbl-cap: "Contents and provenance counts, read from current registers. Source: tables/T-build-report.rds."
#| echo: false
knitr::kable(bld)
```

Three rows carry the discipline the rest of the book rests on.

**`r val("Numbered results")` numbered results, of which `r val("  of which derived-here")`
are `derived-here`.** The convention (@sec-ch02) is that this tag asserts no priority,
and V7 applies a lexical guard to the prose around each such result. @sec-appendix-a
records the proof status and location for every numbered result.

**The `r val("Bibliography entries")` bibliography entries include
`r val("  of which primary-read (independent)")` independent `primary-read` sources
and `r val("  of which programme-artifact-read")` `programme-artifact-read`
sources.** Both tiers require a bounded receipt, but they carry different authority.
The first records a locator read in the independent literature. The second records a
direct inspection of this research programme's own book, manuscript, package, code, or
frozen table and establishes fidelity to that artifact, not independent corroboration.

**`r val("Corrections to the manuscript")` corrections, `r val("  of which carry an author ruling")`
with an author ruling.** @sec-appendix-c is generated from the same correction register.

## All nine verdicts come from this run {#sec-appg-verifiers}

```{r}
#| label: tbl-verifiers
#| tbl-cap: "V1--V9 current outputs and verdict counts. Every underlying row in T-verifiers carries the single run token printed above. Source: tables/T-verifiers.rds."
#| echo: false
knitr::kable(ver_view)
```

`code/R/14-build-all.R` is the release driver. At build start it removes only direct
regular entries under this repository's `verification/` directory and refuses to
continue if that target resolves elsewhere or contains a directory that would require
recursive deletion. It then creates a new token. Consequently, a verifier cannot be
satisfied by a prior run's output.

V1 and V2 share `code/R/03-citation-check.R`, but their current output contracts are
different. V1 writes `citation-check.csv`, `result-check.csv`, the legacy-named
`part-vi-source-receipt-check.csv` (now populated from every registered part receipt),
`annotation-evidence-check.csv`, and `narrative-citation-check.csv`. The annotation
check binds the separate programme-artifact and bounded narrative receipt registers;
it does not collapse programme evidence into independent literature. V2 currently has
no standalone issue CSV; its receipt is the captured `v1-v2-citation-anchor.log`, whose
unique `parity issues` verdict is parsed by the report. The obsolete
`narrative-citation-audit-check.csv` and `result-anchor-check.csv` names are neither read
nor receipted.

The enforced capability of each verifier is narrower than its label:

| | Blocking capability |
|---|---|
| **V1** | Missing bibliography keys; invalid result provenance/source locators; required semantic-receipt coverage; separate programme-artifact and bounded narrative receipt schemas/authority; held-source read-claim violations; and the current narrative-citation tier constraints |
| **V2** | Text/register anchor parity, uniqueness, and chapter/file agreement, reported by the combined V1/V2 run log |
| **V3** | The registered lexical domain of Greek/styled symbols, bare uppercase Latin symbols, indexed lowercase Latin symbols, and exact collision targets; ordinary local lowercase calculus dummies remain outside its scope |
| **V4** | Registered cross-book boundary mentions, real heading anchors outside fenced code, and declared boundary contracts |
| **V5** | Reviewer-inventory/claim-map coverage and valid evidence anchors for addressed or partially addressed rows |
| **V6** | Rendered cross-references and citations, result anchors, manual-number duplication, figure/sidecar/generator bindings, required generated artifacts, selected front-matter count parity, and source-newer-than-render staleness |
| **V7** | Surviving verification markers and lexical novelty language around results registered as `derived-here`; it is not a prior-art search |
| **V8** | The registered numerical checks plus coverage of every `derived-here` result by at least one mapped check |
| **V9** | Required specification artifacts, appendix/answer bindings, and recorded departures; its contents-keyword coverage remains advisory rather than semantic proof |

## Four claim-level checkers sit outside V1--V9 {#sec-appg-claim-checks}

The nine verifiers audit the repository: keys resolve, anchors match, registers
parse, artifacts exist. **None of them reads a sentence and asks whether the
number in it is true.** An external review of this edition found exactly that gap, and the
checkers below were added to close the part of it that can be automated.

```{r}
#| label: tbl-claim-checks
#| tbl-cap: "Claim-level checkers added after external review, with their current outputs and verdict counts. An unrejected negative fixture counts as an issue. Source: tables/T-claim-checks.rds."
#| echo: false
knitr::kable(claim_checks)
```

The contract they enforce is narrower than the phrase *every number is checked*
would suggest, and stating it precisely matters more than stating it warmly.
`r val("Audited evidence claims")` claims imported from the companion volumes
are entered in `manifest/evidence-claim-register.csv` with their chapter, line,
source edition, snapshot locator, and field. Of those,
`r val("  of which independently recomputed")` are recomputed from the frozen
store at build time and fail the build on divergence; the remainder are
**registered but not executed** — their source and value are on the record and
auditable by hand, which is a weaker guarantee and is labelled as one in the
register's own `assertion_mode` column. `r val("Negative fixtures rejected")`
negative fixtures confirm that the executable half rejects the mutations it is
supposed to reject rather than passing everything put in front of it.

Two limits follow, and a reader should hold both. Prose sentences in Part IX are
written by hand, not generated from the store, so the register is what binds them
to a source, not the rendering pipeline. And a registered-but-not-executed row
certifies that someone looked, not that a machine re-derived.

## How the release sequence closes the stale-report loop {#sec-appg-method}

The driver first regenerates bibliography, tables, figures, and non-V1--V9 evidence
checks, then performs a bootstrap Quarto render so V6 has a complete rendered tree to
inspect. V1--V9 run in order. A verifier receives its current-run receipt only after a
zero exit and after every output in its explicit contract exists; the receipt records
the script, filename, token, byte count, SHA-256, completion time, and exit status.

Only after all nine verifier receipts exist does `code/R/16b-build-report.R` generate
`T-build-report` and `T-verifiers`. Quarto then renders the current-run report. Because
that render changes the HTML tree, V6 runs once more against the final render and its
receipt is replaced under the same token. A validation-only report pass recomputes both
tables and requires exact equality with the frozen rendered inputs without rewriting
them and making the render stale again.

This sequence is the release gate implemented by `code/R/14-build-all.R`; it is not a
restriction on Quarto itself. A user can still run `quarto render book` directly, but
that command alone creates no current-run receipt and is not a certified release build.

## Numerical checks are corroboration, not proof {#sec-appg-checks}

V8 ran `r val("Numerical checks (V8)")` checks and recorded
`r val("  of which failed")` failed calculations. Its separate coverage output also
requires every result registered as `derived-here` to map to at least one nonblank
check ID.

These checks corroborate implementations and worked algebra. They are not proofs.
@sec-appendix-a records mathematical arguments and source status; a numerical check can
still encode the same misunderstanding as the prose it was intended to test.

## Reproducing a certified release {#sec-appg-repro}

```bash
Rscript code/R/14-build-all.R
```

The command intentionally replaces prior regular files under `verification/`, creates
a fresh run token, regenerates the current outputs, performs both renders and the final
V6 pass, and finishes with receipt/hash validation. Session information and the run
token are written to `verification/session-info.txt`; the normalized receipt ledger is
`verification/verifier-run-receipts.csv`.

## Limits and maintenance rule {#sec-appg-limits}

This machinery does not certify that the book is substantively correct, that every
citation says what the prose claims, that the lexical screens cover every possible
wording, or that every prose number is generated. It certifies the narrower properties
listed above and that the displayed V1--V9 verdicts come from the exact files receipted
for one build run. Source reading and external review remain necessary; the project's
review record, held with the working repository rather than the published release,
carries that evidence.

When a verifier begins or stops writing an output, both explicit contracts in
`code/R/14-build-all.R` and `code/R/16b-build-report.R` must change together. The next
release must then pass the full driver and all three stale/missing-output mutation
fixtures. This duplicated contract is deliberate: the producer and the report consumer
must agree before Appendix G can be regenerated.
