# Notation, Terminology, and Two Frameworks {#sec-ch02}

```{r}
#| label: setup-ch02
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

A reader who opens this book at chapter 8 and finds $\bar w$ needs to know what it
means, and a reader who moves between this book and the manuscript it stands behind
needs to know where the two differ. This chapter fixes both. It also settles a
vocabulary dispute that turns out not to be about vocabulary: whether the assumption
$\theta_p \sim G$ is a *prior* or a *population distribution*, and what follows from
the answer.

## The data {#sec-ch02-data}

A test of $I$ items is administered to $P$ persons. Write $p = 1,\dots,P$ for persons
and $i = 1,\dots,I$ for items. The response of person $p$ to item $i$ is a random
variable $U_{pi} \in \{0,1\}$, with $1$ denoting a correct response or an endorsement;
its realization is $u_{pi}$. Capitals denote random variables and lowercase their
realizations, and the distinction is held throughout: it is the difference between
$\operatorname{Var}(U_{pi})$, a property of the model, and the observed $u_{pi}$, a
number.

Two collections of responses matter, and they are not interchangeable. The
**within-person response vector**

$$
\mathbf{u}_p = (u_{p1}, \dots, u_{pI})^\top \in \{0,1\}^I
$$

collects one person's responses across items, and it is the object every ability
estimator in this book takes as input. The **item response vector**
$\mathbf{u}^i = (u_{1i},\dots,u_{Pi})^\top$ collects one item's responses across
persons, and it is what item calibration consumes. Their margins are the **person total
score** $r_p = \sum_i u_{pi}$ and the **item total score** $c_i = \sum_p u_{pi}$. Whether
$r_p$ carries all the information about $\theta_p$ that the data contain is not a
notational question but a modelling one, and @sec-ch03 shows that the answer separates
the Rasch model from the 2PL.

::: {.callout-note appearance="simple"}
::: {.ms}
**What the submitted manuscript says.** The manuscript writes $\mathbf{u}_p$ in its
central equations without ever defining it, which Reviewer 2 flagged. It also writes
responses as $y_{ip}$, with the item index first. This book uses $u_{pi}$, person index
first, matching the person-major layout of a response matrix and the convention of the
appendix drafts that seed @sec-ch03 through @sec-ch11.
:::
:::

## Notation {#sec-ch02-notation}

@tbl-notation is the master table. It is generated from
`manifest/notation-register.csv`, which the build lints every chapter against: a symbol
that appears in a numbered equation but not in the register fails check V3.

```{r}
#| label: tbl-notation
#| tbl-cap: "Master notation. Source: tables/T-notation.rds."
#| echo: false
knitr::kable(tbl("T-notation"))
```

Four choices in that table depart from the manuscript, and each is deliberate.

**$P$ for persons, $N$ for the design.** The manuscript uses $N$ for the number of
examinees. This book reserves $N$ for the simulation-design sample size that the
companion volume varies, and writes $P$ for the index bound, so that a statement made
across the two books is unambiguous about which quantity it means. In a realized design
cell with no exclusions or missing persons, $P=N$; $N$ names the planned design factor
and $P$ names the number of response vectors actually analyzed.

**One symbol for information.** Information is $\mathcal{J}(\theta)$ everywhere. The
manuscript's text alternates between $\mathcal{J}$ and $\mathcal{I}$ within a single
paragraph, and the appendix drafts use $\mathcal{I}$ exclusively; the simulation-study
book uses $\mathcal{J}$. Cross-book consistency decides it.

**$\boldsymbol\delta = (\mu_\theta, \sigma^2_\theta)^\top$, with an equals sign.** The
manuscript writes $\{\mu_\theta,\sigma^2_\theta\} \in \delta$, which says the
hyperparameters are *elements of* $\boldsymbol\delta$ rather than *being*
$\boldsymbol\delta$. Reviewer 2 was right to correct it.

**Posterior variance is $v_p$, not a person-indexed lambda.** Shen and Louis, whose
estimators @sec-ch18 and @sec-ch19 develop, use the latter convention for the posterior
variance of $\theta_p$. This book carries the 2PL alongside the Rasch model, so
$\lambda_i$ is the item discrimination on most of the pages where that source symbol
would appear. Rather than
subscript-disambiguate a Greek letter across two hundred pages, the posterior variance
becomes $v_p$. The departure is restated at its first use in @sec-ch11 and @sec-ch18 and
recorded in the crosswalk of @sec-appendix-b.

**$\bar w$ is a reliability coefficient, not the mean of $w_p$.** The overbar is
historical notation inherited from the manuscript and Wright and Masters, not an
averaging operator. This book uses $P^{-1}\sum_p w_p$ when it means the average
shrinkage weight. The two quantities coincide only under constant conditional error;
@sec-ch08 states the approximation and its Jensen direction.

## Two models, and what separates them {#sec-ch02-models}

The book carries the Rasch model and the two-parameter logistic model together. Writing
$\pi_{pi} = \Pr(U_{pi} = 1 \mid \theta_p, \beta_i, \lambda_i)$, the 2PL is

$$
\operatorname{logit}(\pi_{pi}) = \lambda_i(\theta_p - \beta_i),
$$

and the Rasch model is the special case $\lambda_i \equiv 1$. Almost everything in this
book holds for both. Three things do not, and they are worth naming here because they
organize several later chapters:

1. **Sufficiency.** Under the Rasch model $r_p$ is sufficient for $\theta_p$. Under the
   2PL weighted score $\sum_i \lambda_i u_{pi}$ is sufficient only conditional on fixed,
   known discriminations. With unknown $\boldsymbol\lambda$ it is parameter-indexed and
   is not an observed statistic available for Rasch-style conditioning (@sec-ch03).
2. **Conditional maximum likelihood.** Ordinary Rasch CML conditions on person totals to
   eliminate the person parameters. The 2PL has no analogous parameter-free total-score
   reduction, so this exact Rasch construction does not carry over (@sec-ch05). This
   matters because Rasch CML assumes nothing about $G$ and therefore provides a reference
   point for measuring the effect of a distributional assumption.
3. **Scale.** The Rasch model has a location indeterminacy and no scale indeterminacy;
   the 2PL has both. The 2PL therefore needs two identifying constraints where the Rasch
   model needs one, and the second constraint pins the scale of $G$ itself (@sec-ch04,
   @sec-ch16).

These three properties make the Rasch model useful for isolating the effect of assuming
a shape for $G$. They are substantive restrictions, not an appeal to tradition.

## A distributional assumption is not the same thing as a prior {#sec-ch02-frameworks}

The manuscript's exposition moves from maximum likelihood to Bayesian estimation without
saying why, and in doing so leaves the impression that assuming $\theta_p \sim
N(\mu_\theta,\sigma^2_\theta)$ is something Bayesians do and frequentists do not.
Reviewer 2 objected, and the objection is correct.

Joint maximum likelihood treats the $\theta_p$ as fixed incidental parameters, and
Rasch CML conditions them away; neither method needs a population distribution $G$.
Marginal maximum likelihood and hierarchical Bayes do require one. In MML the person
parameters are integrated out against $G$,

$$
L_M(\boldsymbol\beta, G \mid \mathbf{u}) = \prod_{p=1}^{P} \int
  L(\theta, \boldsymbol\beta \mid \mathbf{u}_p)\, dG(\theta),
$$ {#eq-mml}

and without that assumption this marginal likelihood is undefined. Within MML the
assumption is not optional and it is not Bayesian; it is what makes that formulation
work. What *is* distinctively Bayesian in the comparison below is the second assumption
— a prior on the item parameters. Under this book's item-centering convention it is
implemented through iid Gaussian auxiliaries that are mean-centered to form
$\boldsymbol\beta$; the resulting constrained difficulties are dependent. MML does not
make this prior assumption, treating $\boldsymbol\beta$ as fixed structural parameters.

@tbl-frameworks sets the two side by side.

```{r}
#| label: tbl-frameworks
#| tbl-cap: "The role of $G$ in the two frameworks. Source: tables/T-frameworks.rds."
#| echo: false
knitr::kable(tbl("T-frameworks"))
```

Read down the columns and the practical difference is smaller than the rhetorical one.
The functional form of $G$ is identical; the marginal likelihood @eq-mml is the same
integral the Bayesian model computes. What differs is what happens next: MML maximizes
over $(\boldsymbol\beta, \mu_\theta, \sigma^2_\theta)$ and reports point estimates,
while the Bayesian model assigns hyperpriors and integrates. MML is the limiting case of
the hierarchical model as the hyperpriors flatten and the hyperparameter posteriors
concentrate.

The consequence for this book is that **"assuming normality" is a modelling choice, not
a Bayesian sin** — and, more usefully, that the ways of relaxing it run in parallel. The
frequentist route replaces the parametric $G$ with a nonparametric maximum-likelihood
estimate of the mixing distribution; the Bayesian route replaces it with a prior over
distributions. @sec-ch14 traces the first and @sec-ch15 develops the second, and reading
them as two answers to one question is the correct way to read them.

Shrinkage follows from the hierarchy, not from the framework. Under either, the estimate
of $\theta_p$ borrows from the population: with a Gaussian $G$ the posterior mean is a
weighted average of the within-person estimate and $\mu_\theta$, with weight

$$
w_p = \frac{\sigma^2_\theta}{\sigma^2_\theta + \operatorname{se}(\hat\theta_p)^2}.
$$ {#eq-shrinkage-weight}

Frequentist empirical Bayes computes the same weight from MML point estimates.
@sec-ch11 derives @eq-shrinkage-weight and @sec-ch12 places it in the tradition it comes
from; what matters here is only that the hierarchical structure, and not the choice of
framework, is what produces it.

## Levels, not stages {#sec-ch02-levels}

The manuscript describes its model in terms of a "first stage" and a "second stage."
Reviewer 2 called the usage non-standard, and it is worse than non-standard: it is
ambiguous, because the same word is the ordinary name for the steps of a two-step
estimation procedure — calibrate the items, then score the persons. A reader cannot tell
whether "second stage" names a layer of the model or a step of the fitting.

This book writes **levels** and reserves the word for model architecture.

```{r}
#| label: tbl-levels
#| tbl-cap: "Level vocabulary. Source: tables/T-levels.rds."
#| echo: false
knitr::kable(tbl("T-levels"))
```

Under this vocabulary, MML followed by ability estimation is two *stages of estimation*
applied to a model with two *levels*, and saying so is not a quibble — @sec-ch06 shows
that the properties of the resulting ability estimate depend on exactly this
distinction, because the second stage treats a quantity estimated in the first as
though it were known.

## Conventions {#sec-ch02-conventions}

**Numbered results.** Definitions, theorems, propositions, lemmas, and corollaries are
numbered by chapter and carry a provenance tag in the header:

> **Theorem 8.3** (Jensen gap) · *restated from* Lee (2026, Prop. 2)

*Restated* means the result is established in the cited source and reproduced here,
possibly in different notation. *Adapted* means it is established there for a different
setting and modified here, and the modification is stated. *This book's derivation*
means a search for prior art was run and logged and found none — a claim about the
literature, which fails the same way a citation does, and which the reader should
therefore treat as the least secure of the three.

Every numbered result carries a page- or theorem-level locator to its source. A result
cited without one has not been checked, and the build refuses to release with any such
citation outstanding.

**Proof coverage** is claim-level. Short arguments run inline; longer book-supplied
arguments move to @sec-appendix-a; published theorems not reproduced here remain
`source-only` at a precise locator. The Appendix A index distinguishes these from partial
derivations, so a locator is never described as though it were a stand-alone proof.

**Cross-references** to the companion volume are spelled out — *DPMirt Simulation
Study*, ch. 04 § 3 — rather than given as a bare pointer, so that a printed copy of
this book remains usable.

## Sources and provenance {#sec-ch02-sources}

The data-structure notation and the level vocabulary of @sec-ch02-levels follow the
Appendix A draft prepared for the APM revision (§§ A.2, A.4.3, A.5), which in turn
follows @fox_bayesian_2010 and @debelak_introduction_2022 for the hierarchical framing
and @gelman_bayesian_2013 for the level terminology. That draft carries no citation
authority in this book — it is machine-generated and unreconciled — and every statement
taken from it has been re-derived or checked against the source named here.

The parallel between the marginal likelihood and the hierarchical model in
@sec-ch02-frameworks restates @bock_marginal_1981 for the frequentist side and
@fox_bayesian_2010 for the Bayesian, and follows the Appendix D draft (§ D.5) in
setting them against each other. The observation that MML is the flat-hyperprior limit
of the hierarchical model is standard and is stated here without novelty claim.

The notation departures are this book's own decisions, recorded in
`manifest/notation-register.csv` with the collision each resolves.
