# Identification {#sec-ch04}

```{r}
#| label: setup-ch04
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

Reviewer 2 wrote that the Rasch model is not identifiable without constraints, that the
usual $\theta_p \sim N(0,1)$ fixes location and scale, and that a model estimating the
mean and variance of the latent distribution — or replacing it with a Dirichlet process
mixture — needs additional constraints that the manuscript never states. The objection is
correct as an objection. It is also, as this chapter shows, based on a picture of the
problem that the identification literature does not support in one important parametric
case: when the shape of the random-effects distribution is known and only its scale is
unknown, the Rasch scale is **identified from the data**, not fixed by convention. This
does not extend to an unknown semiparametric $G$, whose finite-item identification failure
is more specific than "needs more constraints."

Getting this right matters beyond answering a review. A model that is not identified does
not estimate what a paper says it estimates, and a recovered distribution $G$ that is
only partially identified supports only some of the statements one might want to make
about it. @sec-ch16 returns to the semiparametric case with the Dirichlet process in
hand; this chapter establishes what identification is and settles the parametric models.

## What identification is {#sec-ch04-what}

A statistical model attaches to each parameter value $\boldsymbol\alpha \in A$ a sampling
distribution $P\{\cdot\,;\boldsymbol\alpha\}$ over the observable data.

::: {#def-ident}
**Identification.** *Restated from* San Martín (2016, § 8.2.2)

The parameter $\boldsymbol\alpha$ is **identified by the observations** if the map
$\boldsymbol\alpha \mapsto P\{\cdot\,;\boldsymbol\alpha\}$ is injective: for
$\boldsymbol\alpha, \boldsymbol\alpha' \in A$,

$$
P\{E;\boldsymbol\alpha\} = P\{E;\boldsymbol\alpha'\}
\ \text{ for every event } E
\quad\Longrightarrow\quad
\boldsymbol\alpha = \boldsymbol\alpha' .
$$ {#eq-identification}
:::

Identification is a property of the model rather than of a realized sample.
@eq-identification quantifies
over sampling distributions, not over data sets. San Martín, following Koopmans, puts
the point sharply: exact knowledge of the family of sampling distributions cannot be
derived from any finite number of observations, and is "the limit approachable but not
attainable by extended observation." Hypothesising that knowledge anyway is what
separates the problems of statistical inference, which arise from the variability of
finite samples, from the problems of identification, which are the limits inference
cannot pass *even with infinite data*.

There is always some function of the
parameters that indexes the sampling distributions one-to-one; the identification
question is whether the parameters one wants are a bijective function of it. This is not
a technicality but the method of proof: San Martín's results are obtained by
constructing the identified parametrization first and then asking what restrictions make
the parameters of interest recoverable from it. The consequence is stated in his
discussion and is worth carrying: the restrictions so obtained are **necessary as well as
sufficient**, which is more than a convention-based argument can deliver.

## Location {#sec-ch04-location}

Take the Rasch model @eq-rasch, and add a constant to every person and every item.

::: {#prp-location}
**Location indeterminacy.** *Derived here; no originality claim*

For any $c \in \mathbb{R}$, the parameter values $(\theta_p + c, \beta_i + c)$ and
$(\theta_p, \beta_i)$ induce the same sampling distribution.
:::

*Proof.* $(\theta_p + c) - (\beta_i + c) = \theta_p - \beta_i$, so every $\pi_{pi}$ is
unchanged; by local independence the distribution of $\mathbf{U}$ is a function of the
$\pi_{pi}$ alone. ∎

The proof is one line and the content is not: the data cannot distinguish "everyone is
one logit more able and every item one logit harder" from the original, because the
model only ever speaks about differences. Fixing the origin is a choice of unit, like
choosing where zero sits on a temperature scale.

## Scale {#sec-ch04-scale}

The Rasch and 2PL models differ in their scale orbit, which determines the restrictions
needed in the rest of Part II.

::: {#prp-scale}
**Scale indeterminacy in the 2PL, and its absence in the Rasch model.**
*Derived here; no originality claim*

Under the 2PL, for any $c > 0$ the values $(c\theta_p,\, \lambda_i/c,\, c\beta_i)$ induce
the same sampling distribution as $(\theta_p, \lambda_i, \beta_i)$. Under the Rasch
model, where $\lambda_i \equiv 1$ is fixed, no such transformation exists.
:::

*Proof.* $(\lambda_i/c)(c\theta_p - c\beta_i) = \lambda_i(\theta_p - \beta_i)$. Under the
Rasch model the same rescaling would require $\lambda_i \mapsto 1/c$, which leaves the
model unless $c = 1$. ∎

So the 2PL has a two-dimensional indeterminacy and the Rasch model a one-dimensional
one. The 2PL needs two restrictions where the Rasch model needs one, and — this is the
part that matters later — the second restriction is what pins the *scale on which $G$
lives*. Under the Rasch model the unit is fixed by the model itself, which is why the
variance of $G$ can be a free parameter with a meaning; under the 2PL a convention has to
supply it. @sec-ch16 shows what that costs when $G$ is nonparametric.

## What actually needs restricting, and when {#sec-ch04-table}

The usual account — "we set $\theta_p \sim N(0,1)$ to fix location and scale" — treats
the two indeterminacies as a single problem to be solved by fiat. The identification
results do not support that reading. What must be restricted depends on **how the person
parameters are specified**, and in one case of practical importance the answer is
*nothing*.

```{r}
#| label: tbl-identification
#| tbl-cap: "What must be restricted, by specification and model. Source: tables/T-identification.rds."
#| echo: false
knitr::kable(tbl("T-identification"))
```

Read the rows in order.

**Fixed effects.** The $\theta_p$ are $P$ unknown parameters. The identified
parametrization is $\{F(\theta_p - \beta_i)\}$, identified because each $U_{pi}$ is
Bernoulli and the $U_{pi}$ are mutually independent; the parameters of interest are
recovered from it once one item is used as the origin. The Rasch model needs
$\beta_1 = 0$; the 2PL needs $\beta_1 = 0$ and $\lambda_1 = 1$.

**Random effects, a known shape $G$ up to scale.** Here $\theta_p=\sigma X_p$, where the
$X_p$ are i.i.d. from a fixed, known distribution $G$ with its origin and unit fixed, and
the parameters of interest are $(\boldsymbol\beta,\sigma)$. Under the regularity
conditions in San Martín (2016, § 8.5.1), the Rasch model needs no additional restriction.
For each item, the marginal probability $\delta_i=\Pr\{U_{pi}=1\}$ is identified, and

$$
p(\sigma,\beta)=\int F(\sigma x-\beta)\,G(dx)
$$

is continuous and strictly decreasing in $\beta$. Thus $\beta_i$ is determined by
$(\sigma,\delta_i)$. Substitution into the bivariate probability
$\delta_{12}=\Pr\{U_{p1}=1,U_{p2}=1\}$ produces a map of $\sigma$ that San Martín proves
is continuous and **strictly increasing** when $I\ge2$. Its inverse identifies $\sigma$,
after which the item difficulties are identified.

This is a theorem about a fixed known shape with one unknown scale parameter. It does not
identify the variance of an arbitrary unknown $G$, and it does not justify treating a
Dirichlet-process-mixture variance as data-identified.

**Random effects, $G$ with free location and scale.** Now $\mu$ is free as well, and one
restriction is needed to place the origin: $\beta_1 = 0$ for the Rasch model,
$\beta_1 = 0$ and $\lambda_1 = 1$ for the 2PL, which additionally requires $I \ge 3$.

::: {.callout-note appearance="simple"}
::: {.ms}
**What the submitted manuscript says.** The manuscript does not discuss identification,
and Reviewer 2 is right that it should. But the reviewer's own framing — that
$\theta_p \sim N(0,1)$ is "typically used to fix the location and scale of the latent
trait", so that estimating $\mu_\theta$ and $\sigma^2_\theta$ leaves the model
unidentified — is also not what the results say.

Setting $\sigma^2_\theta=1$ in the known-shape random-effects Rasch family does not remove
an indeterminacy; in that family $\sigma$ is identified by the strictly monotone
bivariate-probability map. It instead restricts an identified parameter and can
misspecify the family. For an unknown semiparametric $G$, however, finite-item data do not
identify the full distribution or its variance. The revision must state which of these
two models it means before making a scale claim.
:::
:::

## Location constraints act on different models {#sec-ch04-constraints}

Several location conventions are used, but they do not all constrain the same statistical
object. The fixed-effects likelihood contains realized person parameters; the marginal
random-effects likelihood contains their distribution $G$ instead.

::: {#thm-equiv-fixed}
**Fixed-effects location orbit.** *Derived here; no originality claim*

In the fixed-effects Rasch likelihood, the transformations
$(\boldsymbol\theta,\boldsymbol\beta)\mapsto
(\boldsymbol\theta+c\mathbf1_P,\boldsymbol\beta+c\mathbf1_I)$ form a location orbit.
Each of $\beta_1=0$, $\sum_i\beta_i=0$, and $\sum_p\theta_p=0$ selects exactly one member
of that orbit. The selected likelihood, fitted probabilities, and within-person or
within-item differences are invariant across the three representatives.
:::

*Proof.* Substituting the orbit into each constraint gives one linear equation in $c$
with a unique solution. @prp-location makes the likelihood constant on the orbit, and
all within-type differences cancel $c$. ∎

::: {#thm-equiv-random}
**Random-effects location orbit.** *Derived here; no originality claim*

Let $G_c$ be the law of $\Theta+c$ when $\Theta\sim G$. In the marginal Rasch model,
$(\boldsymbol\beta,G)$ and $(\boldsymbol\beta+c\mathbf1_I,G_c)$ induce the same response
distribution. An item anchor, item centering, or a finite shift-equivariant location
constraint on $G$ selects one representative. The realized constraint
$\sum_p\theta_p=0$ is not a constraint on this marginal parameter orbit.
:::

*Proof.* After the change of variable $t=\theta+c$, every term depends on
$t-(\beta_i+c)=\theta-\beta_i$, so the marginal response probabilities are unchanged.
The first three constraints again determine one $c$ when their location functionals
exist. The realized sum is a function of latent draws, not of $(\boldsymbol\beta,G)$, and
therefore does not select a marginal-model parameter. ∎

These likelihood equivalences do not make prior or posterior models interchangeable. A
prior must either be transformed with the orbit or deliberately select a representative;
holding a different prior fixed can change posterior geometry and summaries even when
the likelihood quotient is the same.

**Interpretation of the recovered $G$.** Fixing $\mu_\theta = 0$ makes the person
distribution centred *by construction*, so its estimated mean carries no information.
Fixing $\sum_i \beta_i = 0$ leaves $\mu_\theta$ free and estimable, and it is then a
statement about where this population sits relative to this item set. If the recovered
distribution is the object of interest — and in this book it usually is — the second is
the informative choice.

**Posterior geometry.** A sum-to-zero constraint on items is compatible with a free
population location. Imposing a realized person-centering constraint inside a model that
also assigns a population distribution changes the latent-draw model and can create a
redundancy with its location hyperparameter.

A constraint has to be representable in the model as specified. A Dirichlet process
prior on $G$ cannot be made to satisfy
$\sum_p \theta_p = 0$, because the $\theta_p$ are exchangeable draws and the constraint
is a property of the realized set. @sec-ch16 makes this precise; it is the reason the
companion simulation constrains items rather than abilities.

## A proper posterior is not identification {#sec-ch04-bayes}

If the joint prior is proper and the likelihood is integrable under it, the posterior is
proper even when the likelihood has an unidentified direction. It may have a mean, a
variance, and a credible interval, and none of that establishes identification. Conditional
on an identified likelihood orbit, the posterior distribution along the orbit is
proportional to the corresponding conditional prior because the likelihood is constant
there. The data may still inform identified combinations, so the marginal posterior of an
unidentified coordinate need not literally equal its marginal prior when the prior couples
the two directions.

Diagnostics do not catch this problem reliably: a
model with a proper prior on an unidentified direction can converge cleanly. Moreover, the
conditional credible interval along an unidentified orbit is prior-driven; it should not
be presented as information supplied by the observations.

The defensible options are to impose a constraint that selects an identified
representative, or to sample a deliberately expanded model and report only invariant,
identified functionals after rescaling. Both require the transformation, prior, and
reported estimand to be explicit. A proper prior can regularize or select a convention;
it cannot by itself prove that the likelihood identifies the selected coordinate.

## Two results that should be better known {#sec-ch04-heuristics}

San Martín's discussion closes with two remarks that bear directly on how this literature
is usually written, and both are stated here because the manuscript, like most papers,
implicitly assumes the opposite of the first.

**Fixed-effects identification does not imply random-effects identification.** There is a
heuristic in the psychometric literature that it does — San Martín attributes the rule to
Adams et al. (1997) and Adams and Wu (2007) — and the results summarized in
@tbl-identification refute it: the two specifications require different restrictions,
and neither set follows from the other. By extension the heuristic fails for
semiparametric versions too. Since almost every applied paper reasons about
identification informally, and informally usually means by this heuristic, the point
deserves more circulation than it has.

**Identification analysis presumes the model is true, which makes it a design
requirement.** An identification restriction is not only a statement about parameters; it
is sometimes a statement about what the test must contain. San Martín's example is the
1PL-G model, whose identification requires a guessing parameter fixed at zero — which
means the instrument must include an item nobody can get right by guessing. A model
should be applied to data that, *by design*, satisfies its identification restrictions,
and that is a fact about test construction rather than about estimation.

## What is coming, and what is open {#sec-ch04-forward}

The semiparametric case — $G$ itself an unknown parameter — is the one this book's
subject requires, and it is deferred to @sec-ch16 because it needs the Dirichlet process
first. Under the anchored Rasch model, the item parameters and $I+1$ response-score
functionals of $G$ are identified from the response-pattern distribution, but the full
$G$ is **not**. Its variance is not generally among the identified finite-item
functionals, so later reliability claims cannot silently substitute a recovered DPM
variance for an identified population quantity.

One caveat belongs here rather than there, because it qualifies this book's whole scope.
Under TD-3 this book carries the 2PL, and the companion simulation fits Dirichlet process
mixtures under both models. **The identification of the semiparametric 2PL is an open
problem**; San Martín's Table 8.1 marks it with question marks and his discussion says so
directly. Accordingly, every DPM×2PL analysis in this project must pre-specify its location
and scale constraints, label distributional recovery as convention-dependent and
exploratory, and report sensitivity to alternative valid rescalings. Reliability must be
defined as an invariant response-distribution functional or explicitly labelled as a
convention-dependent parameter summary. No 2PL DPM result is called identified unless a
separate theorem establishes it.

@sec-ch26 returns to this boundary after the Rasch semiparametric result is in hand. It
states the 2PL location-scale orbit, the software's constraint pair, and exactly why the
finite-item Rasch theorem is not extended by analogy.

## Sources and provenance {#sec-ch04-sources}

This chapter is built on San Martín's [-@sanmartin_identification_2016] chapter in the
*Handbook of Item Response Theory*, which is the reference treatment: @def-ident is his
§ 8.2.2, the fixed-effects results are § 8.4, the random-effects results including the
$\sigma$-from-bivariate-probabilities argument are § 8.5.1, @tbl-identification condenses
his Table 8.1, and the two remarks of @sec-ch04-heuristics are his § 8.7. The
epistemological framing of identification as a limit not attainable by extended
observation is Koopmans's, quoted there.

@prp-location, @prp-scale, @thm-equiv-fixed, and @thm-equiv-random are elementary orbit
arguments derived here without originality claims. San Martín establishes identification
under specific fixed- and random-effects restrictions; he does not state the former
four-way theorem that an earlier version of this chapter attributed to him. The split
above keeps conditional fixed-effects likelihood, marginal random-effects likelihood,
and prior/posterior models distinct.

The semiparametric results previewed in @sec-ch04-forward are San Martín et al.
[-@sanmartin_bayesian_2011], reported in San Martín (2016, § 8.6) as his Theorems 8.4 and
8.5; @sec-ch16 takes them from the primary source.

@sec-ch04-bayes is standard and is stated without novelty claim. Its claims are
conditional on a proper joint prior and a finite prior predictive normalizing constant;
the orbit statement is about the conditional prior along a flat likelihood direction,
not an unconditional identity between marginal prior and posterior distributions.
