# Estimating Item Parameters {#sec-ch05}

```{r}
#| label: setup-ch05
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

Reviewer 1 objected that the manuscript switches from maximum likelihood to Bayesian
estimation without saying why, and particularly that it gives the item difficulties a
prior without explanation. The objection is fair, and the answer is not a preference. It
is that the person parameters are a nuisance whose number grows with the sample. They can
be maximized over, conditioned away, or integrated out; splitting integration by how $G$
is specified and whether priors are assigned yields the five approaches compared here.

This chapter sets out the four, and it does so with an eye on a question the next
chapters need answered: **what does each method assume about $G$?** That is the axis along
which they differ most, and it is the axis this book cares about.

## The incidental-parameters problem {#sec-ch05-incidental}

Write the joint likelihood over all persons and items,

$$
L(\boldsymbol\theta, \boldsymbol\beta \mid \mathbf{u})
  = \prod_{p=1}^{P}\prod_{i=1}^{I} \pi_{pi}^{u_{pi}}(1-\pi_{pi})^{1-u_{pi}},
$$ {#eq-joint-lik}

and maximize it over everything at once. This is **joint maximum likelihood** (JML), and
it is the obvious thing to do. It does not work, and the reason is structural rather than
computational.

The item parameters are **structural**: there are $I$ of them and $I$ stays fixed as more
people take the test. The person parameters are **incidental**: there is one per person,
so their number grows with the sample. Ordinary maximum-likelihood asymptotics assume the
parameter count is fixed while information accumulates, and here it is not — every new
person brings $I$ new observations and one new parameter, so the information per
parameter never grows.

::: {#thm-neyman-scott}
**Inconsistency of joint maximum likelihood.** *Restated from* Haberman
(2016, § 9.2.1), reporting Andersen (1973, pp. 66–69); the general phenomenon is Neyman
and Scott's (1948)

Under the Rasch model with $I$ fixed and $P \to \infty$, the JML estimator of the item
parameters does not converge to the truth. In the classical two-item case, the estimated
difficulty difference converges almost surely to **twice** its true value.
:::

The two-item case is worth carrying because it makes the failure concrete: the estimator
is not noisy, it is systematically wrong by a factor, and no amount of additional data
fixes it. The familiar $(I-1)/I$ correction applied in practice is an attempt to undo
this factor and is a patch on a structural problem rather than a solution to it.

Haberman adds a qualification that the usual telling omits, and it matters for how
severely one should judge JML. If $I$ grows as well as $P$ — specifically if
$\log(I)/P \to 0$ with the parameters uniformly bounded — then JML **is** consistent for
both the item and the person parameters. The inconsistency is a fixed-$I$ phenomenon.
This theorem does not supply a finite-sample threshold at which the bias becomes small;
claims about a particular 10- or 60-item design require separate simulation or empirical
evidence.

## Conditioning it away {#sec-ch05-cml}

@thm-suff said that under the Rasch model $r_p$ is sufficient for $\theta_p$. Sufficiency
means the conditional distribution of $\mathbf{U}_p$ given $r_p$ does not depend on
$\theta_p$ at all, so conditioning on the total score removes the incidental parameters
exactly rather than approximately:

$$
L_C(\boldsymbol\beta \mid \mathbf{u}, \mathbf{r})
  = \prod_{p=1}^{P} \Pr\{\mathbf{U}_p = \mathbf{u}_p \mid r_p; \boldsymbol\beta\}.
$$ {#eq-cml}

Maximizing @eq-cml is **conditional maximum likelihood** (CML), and the resulting
estimator of $\boldsymbol\beta$ is consistent as $P \to \infty$ with $I$ fixed.

The property that matters for this book is not consistency but what CML does *not*
require: @eq-cml contains no distribution for $\theta$. The person parameters have not
been integrated out against an assumed $G$; they have been conditioned away. JML also
specifies no $G$, but it is inconsistent in the fixed-$I$ regime. Unrestricted marginal
methods avoid a parametric shape for $G$ but inherit identification qualifications. CML's
distinctive conjunction is **no model for $G$, exact elimination of person parameters,
and fixed-$I$ consistency for Rasch item parameters**. That makes it the clean conditional
reference for measuring the cost of a distributional assumption.

The price is computational and, more importantly, structural. The conditional
probabilities involve elementary symmetric functions of the item parameters, whose
evaluation is delicate for long tests. And @prp-nosuff says the whole construction is
unavailable under the unrestricted 2PL in its ordinary Rasch form: the
parameter-indexed weighted score cannot serve as a parameter-free person total. A 2PL
analysis can still be used as a generalization or sensitivity analysis, but it lacks this
exact Rasch conditional anchor.

## Integrating it away {#sec-ch05-mml}

The third option treats the $\theta_p$ as draws from a distribution and integrates:

$$
L_M(\boldsymbol\beta, G \mid \mathbf{u})
  = \prod_{p=1}^{P} \int L(\theta, \boldsymbol\beta \mid \mathbf{u}_p)\, dG(\theta).
$$ {#eq-mml-ch5}

This is **marginal maximum likelihood** (MML). The incidental parameters are gone, the
method accommodates fixed $I$, and restricted MML extends to an identified parametric
2PL once location and scale constraints are imposed. Consistency additionally requires
correct specification of the chosen family for $G$, identification, and the usual
regularity conditions; it is not a property of the integral alone.

What it costs is stated in @eq-mml-ch5 itself: $G$ appears, and something must be said
about it. Two things can be said, and the distinction is Haberman's.

**Restricted marginal estimation** fixes a parametric family — almost always
$G = N(\mu_\theta, \sigma^2_\theta)$ — and estimates its parameters alongside
$\boldsymbol\beta$. This is what "MML" means in ordinary usage and what the manuscript
does. The integral is then computed by Gauss–Hermite quadrature and the whole maximized by
EM: treating the $\theta_p$ as missing data, the E step forms the posterior of each
person's trait under current parameters and the M step maximizes the expected complete
log-likelihood, which separates by item. That algorithm is Bock and Aitkin's
[-@bock_marginal_1981], and it is what made MML practical.

**Unrestricted marginal estimation** leaves $G$ completely free and maximizes @eq-mml-ch5
over $\boldsymbol\beta$ and over the space of distribution functions jointly. The
maximizer is a discrete distribution on finitely many support points. This is
nonparametric MML, and it is the frequentist counterpart of what Part V does with a prior
over distributions — @sec-ch14 traces it and @sec-ch15 develops the Bayesian route. It is
worth naming here because the two are usually presented as unrelated traditions when they
are two answers to one question. It is not a license for categorical consistency claims:
Haberman warns that unrestricted estimation can be unsatisfactory, and finite-item Rasch
data identify only anchored item parameters and finitely many functionals of $G$, not the
full mixing distribution. For a semiparametric 2PL, identification remains open.

## Giving the items a prior {#sec-ch05-bayes}

The fifth approach is the manuscript's, and it is a Bayesian version of marginalization.
Under this book's preferred sum-to-zero item constraint, independent priors are placed on
unconstrained auxiliaries and then centered:

$$
\widetilde\beta_i \overset{\text{iid}}{\sim} N(0,\sigma^2_\beta),
\qquad
\beta_i=\widetilde\beta_i-\frac1I\sum_{j=1}^I\widetilde\beta_j,
\qquad \sum_i\beta_i=0.
$$ {#eq-item-prior}

The induced prior on $\boldsymbol\beta$ is exchangeable on the constrained subspace but
is not iid: centering makes the difficulties dependent. Hyperpriors are placed on
$\sigma^2_\beta$ and on $G$'s parameters, and the joint posterior is reported. An
anchor-based parameterization could instead put priors on the free item coordinates, but
the two prior constructions should not be conflated.

@sec-ch02 established that this is *not* where the distributional assumption on $\theta$
enters — restricted MML makes that assumption too. @eq-item-prior is the genuinely
additional regularization model, and Reviewer 1 was right to ask what it buys.

**Partial pooling across items.** An item answered correctly by everyone has no finite
unregularized difficulty estimate; under a proper version of @eq-item-prior it has a
finite posterior, shrunk toward the constrained item center. The size of that effect is a
prior-sensitivity question.

**Coherent propagation of uncertainty.** In a two-stage procedure the item estimates are
treated as known when persons are scored, which understates the uncertainty in
$\hat\theta_p$ — @sec-ch06 makes this precise and it is the substance of Reviewer 2's
third bullet. Fitting everything jointly carries the item uncertainty into the person
posterior automatically.

**A route to a flexible $G$.** Once $G$ is a parameter with a prior, replacing that prior
with one over distributions is a change of prior rather than a change of method. That is
what makes the semiparametric extension of Part V natural in this framework and awkward
in the others.

The price is a second modelling assumption. @eq-item-prior is exchangeable over item
labels before centering and regularizes the constrained vector toward a common center.
For a test assembled from a blueprint with deliberate difficulty targets, that is best
treated as a regularization device whose influence is checked, not automatically as a
literal model of item writing.

## Five approaches, side by side {#sec-ch05-compare}

```{r}
#| label: tbl-estimation
#| tbl-cap: "Five approaches to incidental person parameters, with model and identification scope. Source: tables/T-estimation.rds."
#| echo: false
knitr::kable(tbl("T-estimation"))
```

The $G$ column is only one axis. The model/scope and identification columns prevent a
shape-free mixing distribution from being read as an identified full $G$, and they keep
the open semiparametric-2PL problem visible. Part V weakens parametric shape assumptions;
it does not thereby solve identification.

## Sources and provenance {#sec-ch05-sources}

The incidental-parameters framing and @thm-neyman-scott follow Haberman's
[-@haberman_nuisance_2016] chapter, which is the reference treatment: the JML
inconsistency and the two-item example are his § 9.2.1, reporting Andersen (1973,
pp. 66–69), and the consistency result under growing $I$ is his citation of his own
earlier work. The distinction between restricted and unrestricted marginal estimation is
his §§ 9.4–9.5, and the terminology is worth adopting because the usual vocabulary has no
name for the difference. Neyman and Scott [-@neyman_consistent_1948] named the general
phenomenon and are cited through Haberman under rule R3b; the primary is not held.

The EM algorithm for restricted marginal estimation is Bock and Aitkin's
[-@bock_marginal_1981]. CML's development follows Debelak et al.
[-@debelak_introduction_2022, § 3.2], and the sufficiency it rests on is @thm-suff.

The argument of @sec-ch05-bayes is assembled here rather than taken from a source. Its
three components — partial pooling, uncertainty propagation, and the route to a flexible
$G$ — are individually standard; the claim that the third is what makes the Bayesian
framework the natural home for a semiparametric $G$ is this book's framing, and it is
argued rather than asserted in @sec-ch15 and @sec-ch16.
