# Estimating Person Parameters {#sec-ch06}

```{r}
#| label: setup-ch06
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

Reviewer 2 asked for the mathematical expression of $\hat\theta_p$ and pointed out that
the manuscript's claim — that ability can be estimated by maximum likelihood — holds only
when the item parameters are known, whereas practice substitutes estimates. Both points
are correct, and this chapter answers them.

It also does something the usual treatment does not. Maximum likelihood, Bayesian modal
estimation, and Warm's weighted likelihood estimator are normally presented as three
competing methods. They are one estimating equation with three choices of a single
function, and seeing that makes the differences between them exact rather than
impressionistic. The unification is Warm's own, in the paper that introduced the third.

## One equation, three estimators {#sec-ch06-family}

Fix the item parameters for the moment. Write $P_i(\theta)$ for the probability of a
correct response to item $i$, $Q_i = 1 - P_i$, and $P_i' = \partial P_i/\partial\theta$.
The person's likelihood is $L(\mathbf{u}_p \mid \theta) = \prod_i P_i^{u_{pi}}Q_i^{1-u_{pi}}$,
and maximizing it means solving

$$
\frac{\partial \ln L}{\partial\theta}
  = \sum_{i=1}^{I}\frac{(u_{pi} - P_i)P_i'}{P_i Q_i} = 0 .
$$ {#eq-score}

Under the Rasch model $P_i' = P_iQ_i$, so @eq-score collapses to
$r_p = \sum_i P_i(\hat\theta_p)$: **choose the ability at which the expected number
correct equals the observed number correct.** That is the explicit expression Reviewer 2
asked for, and it is worth stating in that form because it makes the estimator's meaning
plain.

Now consider maximizing not the likelihood but $f(\theta)L(\mathbf{u}_p \mid \theta)$ for
some positive function $f$. The estimating equation becomes

$$
\sum_{i=1}^{I}\frac{(u_{pi} - P_i)P_i'}{P_i Q_i}
  \;+\; \frac{\partial \ln f(\theta)}{\partial\theta} = 0 .
$$ {#eq-warm-family}

@eq-warm-family is the whole family, and three choices of $f$ name three estimators.

| $f(\theta)$ | Estimator | What it is |
|---|---|---|
| a positive constant | **ML** | @eq-warm-family reduces to @eq-score |
| a prior density for $\theta$ | **MAP** (Bayesian modal, BME) | $f \cdot L$ is proportional to the posterior |
| $w(\theta)$ with $\partial\ln w/\partial\theta = J/(2\mathcal{J})$ | **WLE** | removes the leading bias term |

So the switch from likelihood to posterior that Reviewer 1 found abrupt is not a switch
of optimization template. It is a choice of $f$ in one equation. Whether $f$ represents
a population prior or a test-specific bias-reducing weight is an inferential distinction
even when the resulting objective has the same algebraic form.

## Where the bias comes from, and how WLE removes it {#sec-ch06-wle}

Lord [-@lord_unbiased_1983] gives the asymptotic bias of the ML ability estimate, of
order $I^{-1}$:

$$
\operatorname{BIAS}(\hat\theta^{\mathrm{ML}}) = \frac{-J}{2\mathcal{J}^2},
\qquad
\mathcal{J} = \sum_i \frac{P_i'^2}{P_iQ_i},
\qquad
J = \sum_i \frac{P_i'P_i''}{P_iQ_i}.
$$ {#eq-lord-bias}

For a prior $f$, the bias of the corresponding estimator picks up one extra term,

$$
\operatorname{BIAS}(\theta^*)
  = \operatorname{BIAS}(\hat\theta^{\mathrm{ML}})
  + \frac{\partial \ln f(\theta)/\partial\theta}{\mathcal{J}},
$$ {#eq-warm-bias}

which for a standard normal prior is $-\theta/\mathcal{J}$ — shrinkage toward zero, visible
as bias. Warm's step is to read @eq-warm-bias as an equation to be solved rather than a
quantity to be reported: set the right-hand side to zero and solve for $f$. Substituting
@eq-lord-bias gives

$$
\frac{\partial \ln w(\theta)}{\partial\theta} = \frac{J}{2\mathcal{J}},
$$ {#eq-warm-weight}

and the resulting estimating equation is

$$
\sum_{i=1}^{I}\frac{(u_{pi} - P_i)P_i'}{P_i Q_i} + \frac{J}{2\mathcal{J}} = 0 .
$$ {#eq-wle}

::: {#thm-warm}
**Warm's bias correction.** *Restated from* Warm (1989, eqs. 6–10,
Theorem, and Appendix)

Assume the item parameters are known; $\theta$ is restricted to a bounded interval; the
item response probabilities are continuous and bounded away from zero and one; five
derivatives exist and are bounded; and test length grows by replication in Warm's stated
asymptotic scheme. Then the estimator solving @eq-wle has
$\operatorname{BIAS}(\hat\theta^{\mathrm{WLE}})=o(I^{-1})$, is asymptotically normal, and
has the same **first-order asymptotic** variance $\mathcal{J}^{-1}+o(I^{-1})$ as ML.
:::

Warm proves the theorem in his appendix under these restrictive conditions and reports
Hinkley's suggestion that the method can extend to consistent estimating equations with a
computable bias. That suggestion is not a replacement for model-specific regularity
checks. The theorem concerns interior replicated-item asymptotics, not finite tests at
their all-correct or all-incorrect boundaries.

For the 1PL and 2PL, WLE is algebraically Jeffreys-MAP although Warm's construction and
interpretation are not Bayesian. For these models Warm shows
$w(\theta)=\mathcal{J}(\theta)^{1/2}$, the Jeffreys weight. Maximizing
$w(\theta)L(\theta)$ therefore has exactly the same mode equation as using that
test-specific Jeffreys weight as a prior. The algebra cannot support a claim that WLE is
“not MAP.” Warm's point is inferential: no population distribution for $\theta$ is
assumed, and $w$ is a function of the administered items. He gives the direct comparison:
two populations taking one test use the same WLE weight, while one population taking two
tests uses different weights—the reverse of how a population prior behaves.

The variance equality is first-order asymptotic rather than a finite-sample identity. Under
@thm-warm's conditions WLE and ML have the same leading variance term. Their finite-test
variances can differ, especially near the scale boundary. WLE deliberately uses only the
within-person likelihood and a test-derived adjustment, not population information about
the person. That makes it the natural contrast to EAP; it does not make every finite-sample
risk component equal.

## The 2PL specialization {#sec-ch06-2pl}

For the 2PL,

$$
P_i(\theta)=\operatorname{logit}^{-1}\{\lambda_i(\theta-\beta_i)\},
\qquad P_i'(\theta)=\lambda_iP_iQ_i,
$$

so the ML score and information are

$$
S_{2\mathrm{PL}}(\theta)=\sum_{i=1}^I\lambda_i(u_{pi}-P_i),
\qquad
\mathcal{J}_{2\mathrm{PL}}(\theta)=\sum_{i=1}^I\lambda_i^2P_iQ_i.
$$ {#eq-2pl-score}

The Warm numerator specializes to

$$
J_{2\mathrm{PL}}(\theta)
=\sum_{i=1}^I\lambda_i^3P_iQ_i(1-2P_i),
$$

and the 2PL WLE solves

$$
\sum_{i=1}^I\lambda_i(u_{pi}-P_i)
+\frac{J_{2\mathrm{PL}}(\theta)}{2\mathcal{J}_{2\mathrm{PL}}(\theta)}=0.
$$ {#eq-2pl-wle}

Thus discrimination weights the score, enters information quadratically, and enters the
Warm bias numerator cubically. When all $\lambda_i=1$, these expressions reduce to the
Rasch equations above. Warm's known-item and asymptotic conditions apply to both.

## Where there is no estimate at all {#sec-ch06-boundary}

Under the Rasch model, @eq-score has no finite solution when $r_p=0$ or $r_p=I$. The same
boundary result holds for the 2PL when every $\lambda_i>0$: an all-correct likelihood
increases with $\theta$, and an all-incorrect likelihood decreases. The ML estimate is
$+\infty$ or $-\infty$.

On a twenty-item screening test in a class of sixty this is not an edge case. It is the
students the screening is for.

The estimators respond differently, and the difference is diagnostic of what each is
doing. MAP is finite at the boundary because the prior supplies curvature the likelihood
lacks. For the ordinary logistic Rasch and 2PL models, the
$\mathcal{J}(\theta)^{1/2}$ weight yields finite WLE roots for extreme response patterns.
ML is not finite, and practices such as bounding the estimate, dropping extreme scores,
or assigning a conventional value are analytic decisions that should be reported. No
software-default claim is needed for the mathematical comparison.

## Plug-in scoring and calibration uncertainty {#sec-ch06-plugin}

Everything so far fixed the item parameters. A common two-stage workflow calibrates
items and then scores persons:

$$
\hat\theta_p = \hat\theta_p(\hat{\boldsymbol\psi}),
$$

substituting the calibration estimates into @eq-score or @eq-wle as though they were the
truth. This is the **plug-in** estimator. The inverse root information
$\mathcal{J}(\hat\theta_p)^{-1/2}$ is a local/asymptotic standard error conditional on the
plugged-in item values. Whether it understates an unconditional standard error, and by how
much, depends on how calibration and scoring samples are related.

::: {#prp-plugin}
**Independent-calibration delta method.** *Derived here; no
originality claim*

Let $\boldsymbol\psi$ collect all calibrated item parameters, let
$S_p(\theta,\boldsymbol\psi)=0$ define a regular ability estimator, and suppose
$\hat{\boldsymbol\psi}$ comes from an independent calibration sample with covariance
$V_\psi$. Write

$$
\mathbf{g}_\psi
=-\left(\frac{\partial S_p}{\partial\theta}\right)^{-1}
  \frac{\partial S_p}{\partial\boldsymbol\psi}.
$$

Then the first-order variance is

$$
\operatorname{Var}\{\hat\theta_p(\hat{\boldsymbol\psi})\}
\approx \mathcal{J}(\theta_p)^{-1}
+\mathbf{g}_\psi^\top V_\psi\mathbf{g}_\psi.
$$ {#eq-plugin-delta}
:::

*Proof.* The implicit-function theorem gives the displayed sensitivity
$\mathbf{g}_\psi$. A first-order Taylor expansion around the true item parameters, followed
by independence of the calibration estimate and scoring response, adds the quadratic
delta-method term to the known-item conditional variance. ∎

For the Rasch ML score, the sensitivity to difficulty $\beta_j$ is
$g_{\beta_j}=P_jQ_j/\mathcal{J}$. The 2PL has sensitivities to both difficulties and
discriminations, so $\boldsymbol\psi=(\boldsymbol\beta,\boldsymbol\lambda)$. If calibration
and scoring samples overlap, the independence step fails and cross-covariance terms can
change the result. A stacked estimating-equation sandwich, a design-respecting bootstrap,
or a joint posterior is then required.

The size of the calibration term cannot be inferred from the number of persons being
scored. It depends on the calibration sample size and design, item information,
calibration/scoring overlap, constraints, and the target estimand. A scoring condition
with $P=20$ signals a possible risk; it does not determine the magnitude or direction of
the error by itself.

**Warm's theorem does not cover this.** @thm-warm is proved, in Warm's words, "assuming
the item parameters are known constants." The $O(I^{-1})$ bias that WLE removes is the
bias of the ideal estimator. Whatever bias the plug-in step introduces is a separate
matter that WLE does not address, and treating a WLE as unbiased in an operational
setting conflates the two.

This is the honest form of Reviewer 2's third bullet. The claim is not that plug-in
estimation is wrong. The claim is that its uncertainty target must be named. Chapters 7
and 8 call inverse-information MSEM and reliability **conditional on the calibrated item
values**. A calibration-aware unconditional reliability replaces that error component
with one that includes @eq-plugin-delta or its overlap-aware analogue; the two coefficients
answer different design questions.

## The four estimators {#sec-ch06-compare}

```{r}
#| label: tbl-estimators
#| tbl-cap: "Person-parameter estimators. Source: tables/T-estimators.rds."
#| echo: false
knitr::kable(tbl("T-estimators"))
```

The last column is the one that matters for Part VI. ML and WLE use only the responses of
the person being scored; MAP and EAP additionally use the population. That is not a defect
of either pair — it is the difference between asking "what does this person's test say
about them" and "what is my best guess about this person." @sec-ch17 shows that the two
questions have different answers because they are different questions, and Part VI builds
estimators for a third and fourth.

## Sources and provenance {#sec-ch06-sources}

The unification of @sec-ch06-family is Warm's [-@warm_weighted_1989], and this chapter
follows his development closely because it is clearer than the usual three-methods
presentation: his equations (3) through (5) give the family, (6) is Lord's
[-@lord_unbiased_1983] bias expression, (7) is Lord's (1984) result for the Bayesian modal
estimator under a standard normal prior, (8) is the conjecture that the bias of the
general member picks up one extra term, and (9) and (10) solve for the weight and state
the estimating equation. @thm-warm is his Theorem, proved in his appendix; the remark on
generality is Hinkley's, reported there.

One step of that chain is worth spelling out, because a reader who opens Lord will not
find @eq-lord-bias on the page. Lord derives the bias in the three-parameter
parameterization, as his equations (27) to (29) on p. 237: $B_1(\hat\theta) = I^{-2}\sum_i
A_i I_i(\phi_i - \tfrac12)$ with $\phi_i = (P_i - c_i)/(1 - c_i)$ and $I_i =
P_i'^2/(P_iQ_i)$, and he notes there that $B_1$ is of order $n^{-1}$ because $I$ is of
order $n$ — which is @eq-lord-bias's order claim in his notation. Setting $c_i = 0$ and
using $P_i' = A_iP_iQ_i$ and $P_i'' = A_i^2P_iQ_i(Q_i - P_i)$ turns his sum into $-J/2$,
so the guessing-free form this chapter displays is Warm's rendering of Lord's result
rather than a formula Lord himself writes. The equivalence was checked by hand against
pp. 236–237, where Lord's (20) also supplies the $\mathcal{J}$ used here.

The observation that $w(\theta)$ coincides with the Jeffreys prior is Warm's, citing
Jeffreys (1961) and Akaike (1978); so is the disclaimer that WLE is nonetheless not
Bayesian in derivation or population interpretation. Algebraically, maximizing the
weighted likelihood is the Jeffreys-weight MAP objective for the 1PL and 2PL. Warm makes
the test dependence explicit through his two-populations/one-test and
one-population/two-tests comparison; the administered-test argument is his.

The score equation and its Rasch simplification follow Baker and Kim
[-@baker_item_2004, ch. 3]. @prp-plugin is a first-order implicit-function and delta-method
derivation given here without an originality claim. It assumes independent calibration
and scoring; the overlap alternatives are stated because a marginal law-of-total-variance
identity alone does not supply the required covariance structure.
