# Reliability in the Rasch Model {#sec-ch08}

```{r}
#| label: setup-ch08
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

@sec-ch01 asserted that "reliability below 0.7" is not interpretable until three further
questions are answered: reliability of what, computed how, and a property of what. This
chapter answers them. It is the longest chapter in Part III because the answers are not
difficult but are routinely skipped, and because the whole simulation design in the
companion book rests on treating one particular coefficient as a design factor.

The organizing fact is @eq-reliability-shape: every coefficient in circulation is the
same ratio, of variance attributable to true differences between persons to that variance
plus measurement error variance. They differ in how the two terms are obtained. Once that
is said clearly, most of the confusion in the area dissolves, and what remains — a
genuine ordering among three population functionals, and a genuine ambiguity about
whether $\sigma^2_\theta$ is estimated or set — becomes visible.

## The classical ratio {#sec-ch08-ctt}

Classical test theory writes an observed score as a true score plus error,

$$
X = T + E, \qquad \operatorname{cov}(T, E) = 0, \qquad \operatorname{E}(E) = 0,
$$ {#eq-ctt}

from which reliability is

$$
\rho_{XX'} = \frac{\operatorname{Var}(T)}{\operatorname{Var}(X)}
           = \frac{\operatorname{Var}(T)}{\operatorname{Var}(T) + \operatorname{Var}(E)} .
$$ {#eq-rho-ctt}

The definition is Spearman's: he observed in 1904 that measurement error attenuates
observed relationships, and introduced the term *reliability coefficient* in 1910, as
Adams [-@adams_reliability_2005] recounts. @eq-rho-ctt reappears in enough different
guises that its ubiquity is easy to mistake for depth. Adams lists five: the correlation
between true and observed scores is $\sqrt{\rho_{XX'}}$; the correlation between two
parallel observed scores is $\rho_{XX'}$; a correlation with an external variable is
attenuated by $\sqrt{\rho_{XX'}}$; the variance of the observed mean as an estimator of
the true mean is inflated by $1/\rho_{XX'}$; and $\operatorname{Var}(T) =
\operatorname{Var}(X)\rho_{XX'}$ tautologically.

The fourth is the one @sec-ch08-population returns to.

$\operatorname{Var}(T)$ is not observable, so classical coefficients bound it rather than
estimate it. Coefficient $\alpha$ [@cronbach_coefficient_1951],

$$
\alpha = \frac{n}{n-1}\left(1 - \frac{\sum_i V_i}{V_t}\right),
$$ {#eq-alpha}

with $V_i$ the item variances and $V_t$ the total-score variance, is a lower bound on
$\rho_{XX'}$ under any model — a result van der Ark [-@ark_standard_2025, p. 1679]
attributes to Lord and Novick (1968, Theorem 4.4.3). Guttman's $\lambda_2$
[@guttman_basis_1945, § 13] is a sharper lower bound using the squared inter-item
covariances, and Guttman's $\lambda_3$ is $\alpha$ [@ark_standard_2025, p. 1686, § 3.7]. Three
things follow that are worth saying flatly, because each is regularly got wrong:
$\alpha$ is a bound and not an estimate; it is a statement about the sum score and not
about a latent variable; and it is not a measure of unidimensionality
[@sijtsma_use_2009].

## The IRT reliability zoo {#sec-ch08-zoo}

@tbl-zoo lays out the coefficients that compete for the name in item response theory. It
is not an inventory for its own sake. Read down the second column and the structure of
the field appears: what separates these coefficients is almost entirely *where the true
variance comes from*.

```{r}
#| label: tbl-zoo
#| tbl-cap: "Reliability coefficients in circulation. Implementation formulas were read from package source (TAM 4.3-25, mirt 1.46.1, IRTsimrel 0.2.0), not from documentation. Source: tables/T-reliability-zoo.rds."
#| echo: false
knitr::kable(tbl("T-reliability-zoo"))
```

Three routes are available, and an estimator determines which one is correct.

**Subtraction.** ML and WLE estimates carry measurement error on top of true variation,
so their variance is inflated. Under the assumptions of @eq-ctt applied to
$\hat\theta_p = \theta_p + \varepsilon_p$, Adams [-@adams_reliability_2005, eq. 5] gives

$$
\operatorname{Var}(\hat\theta) = \sigma^2_\theta + \mathrm{MSEM},
$$ {#eq-overdisp}

so the true variance is recovered by subtracting, and reliability is

$$
1 - \frac{\mathrm{MSEM}}{\operatorname{Var}(\hat\theta)} .
$$ {#eq-rel-subtract}

This is the person-separation reliability of Wright and Stone (1979) as Adams
[-@adams_reliability_2005, eq. 6] presents it, and it is exactly what `TAM::WLErel`
computes.

**Addition.** EAP estimates are shrunk toward the prior mean and their variance is
*deflated*. The identity $\operatorname{Var}(\theta) = \operatorname{Var}[\operatorname{E}
(\theta \mid \mathbf{u})] + \operatorname{E}[\operatorname{Var}(\theta \mid \mathbf{u})]$
— which Adams attributes to Verhelst (2001) in this use — gives

$$
\sigma^2_\theta = \operatorname{Var}(\hat\theta^{\mathrm{EAP}}) + \bar\sigma^2_{\text{post}},
$$ {#eq-underdisp}

so the true variance is recovered by *adding* the mean posterior variance, and
reliability is

$$
\frac{\operatorname{Var}(\hat\theta^{\mathrm{EAP}})}
     {\operatorname{Var}(\hat\theta^{\mathrm{EAP}}) + \bar\sigma^2_{\text{post}}} .
$$ {#eq-rel-add}

Adams remarks that @eq-rel-add is "in some sense the reciprocal" of @eq-rel-subtract, and
that is the whole of it: the two formulas are the same operation performed on opposite
sides of an over- or under-dispersed estimate.

**Assumption.** The third route is the manuscript's, and the simulation's. $\sigma^2_\theta$
is neither subtracted nor added; it is *known*, because the design set it. That is
@sec-ch08-design.

::: {.callout-important}
## The two formulas are one function argument apart, and are not interchangeable

`mirt::empirical_rxx` computes @eq-rel-add by default and @eq-rel-subtract when
`T_as_X = TRUE`. Nothing in the call checks which estimator produced the scores.

For any $\mathrm{MSEM} = m > 0$ and any $\operatorname{Var}(\hat\theta) = v > m$,

$$
\frac{v}{v+m} \;-\; \left(1 - \frac{m}{v}\right) \;=\; \frac{m^2}{v(v+m)} \;>\; 0 ,
$$ {#eq-two-forms}

so @eq-rel-add always returns the larger number. Applying it to ML or WLE scores
therefore overstates reliability, and by @eq-overdisp the overstatement is systematic
rather than incidental. The exact algebra above is the claim used here; a numerical
simulation comparison belongs with the companion book's reproducible experiments.
:::

### When reliability comes out negative {#sec-ch08-negative}

@eq-rel-subtract is negative exactly when $\mathrm{MSEM} > \operatorname{Var}
(\hat\theta)$: when the average error variance exceeds the observed spread of the
estimates. This is not a numerical pathology and not a bug in the software. It is what
subtraction does when the design is too weak for the population — few items, a
homogeneous sample, floor or ceiling effects — and @eq-overdisp is violated in sample
because the sampling error in $\operatorname{Var}(\hat\theta)$ swamps $\sigma^2_\theta$.
A negative value carries information: it says the data cannot distinguish
between-person differences from noise at all.

Two consequences. First, @eq-rel-add cannot go negative, since it is a ratio of
non-negative quantities, so the choice of formula changes not only the value but the
range. Second — and this is the point that returns in @sec-ch08-design — a simulation
that *fixes* $\sigma^2_\theta = 1$ can never produce a negative reliability, because it
never subtracts. The phenomenon is invisible by construction to the design that is
supposed to be studying reliability.

## Reliability is a property of a test in a population {#sec-ch08-population}

@eq-rho-ctt has a numerator that belongs to the persons and a denominator that belongs to
the persons and the instrument together. Adams [-@adams_reliability_2005] puts the
consequence plainly: reliability is a characteristic of both the instrument and the
sample it is applied to, so a very homogeneous sample yields low reliability regardless
of instrument quality, and a very heterogeneous one can make an inaccurate instrument
look reliable.

This is why the phrase "the reliability of the instrument" is a category error, and it is
also the licence for everything the companion simulation does. If reliability were a
property of a test, varying it as a design factor would be varying the test. Because it
is a property of a test-in-a-population, it can be varied by moving either side, and
@sec-ch09 shows which side the manuscript moved.

Adams's larger point is that in large-scale survey work the useful reading of reliability
is its reciprocal, which he names the **measurement design effect**: the factor by which
not observing true abilities inflates the sampling variance of a population estimate. For
the mean, that factor is exactly $1/\rho$, from the fourth of the five equivalences in
@sec-ch08-ctt. Two features of it matter here. It varies by statistic — Adams reports
that in PISA 2003 the design effect for country means was mild while the effect for the
percentage of students at a proficiency level was far larger, and in two systems it more
than doubled the error variance. And it is a statement about *population* estimands, not
individual ones, which is precisely the distinction Part VI is built on.

## Reliability and the mean shrinkage weight {#sec-ch08-shrinkage}

@sec-ch01 introduced the shrinkage weight $w_p = \sigma^2_\theta/(\sigma^2_\theta +
\operatorname{se}(\hat\theta_p)^2)$ from @eq-shrinkage-weight, and the manuscript's
reliability is written $\bar w$. The overbar invites the reading that $\bar w$ is the
average of the $w_p$. It is not, except in a special case, and the direction of the error
is knowable.

::: {#prp-mean-weight}
**The mean weight exceeds the reliability.** *Derived here; no originality claim*

Let $s^2_p$ be working conditional variances with $\mathrm{MSEM} = P^{-1}\sum_p s^2_p$,
and let $w_p = \sigma^2_\theta/(\sigma^2_\theta + s^2_p)$. Then

$$
\frac{1}{P}\sum_{p=1}^{P} w_p \;\ge\; \frac{\sigma^2_\theta}{\sigma^2_\theta + \mathrm{MSEM}}
\;=\; \bar w ,
$$ {#eq-mean-weight}

with equality if and only if the $s^2_p$ are constant across persons.
:::

*Proof.* $x \mapsto \sigma^2_\theta/(\sigma^2_\theta + x)$ is convex on $x > 0$; apply
Jensen's inequality to the empirical distribution of the $s^2_p$. Strict convexity gives
the equality condition. ∎

Two qualifications keep @prp-mean-weight honest. The weights $w_p$ are those of a
*normal–normal* working approximation: if the person likelihood is approximated locally
by $N(\hat\theta_p, s^2_p)$ and the prior is $N(\mu_\theta, \sigma^2_\theta)$, the
approximate posterior mean is $w_p\hat\theta_p + (1-w_p)\mu_\theta$. An exact Rasch EAP
is a posterior integral with no such closed form, so @prp-mean-weight is an exact
statement about the working weights and an approximate one about EAP shrinkage. And the
equality condition is not idle: $\operatorname{se}(\hat\theta_p)$ is constant only if
information is constant across the persons measured, which @sec-ch07-information showed
it is not.

The practical reading is that $\bar w$ is a *lower bound* on how much weight a
shrinkage estimator places on the data, and the gap widens as the error curve spreads.
@sec-ch08-functionals shows the same mechanism operating one level up, on population
functionals rather than person averages, and the two together are the reason this chapter
treats Jensen's inequality as structural rather than as a technicality.

### Estimated or set: what $\sigma^2_\theta$ is {#sec-ch08-design}

There is a difference between two operations that produce the same symbol.

Wright and Masters [-@wright_rating_1982, § 5.5] obtain the true variance by
**subtraction**: they define the adjusted person variance as $SA_P^2 = SD_P^2 - MSE_P$,
where $SD_P^2$ is the observed variance of the estimates. The decomposition is therefore
an identity by construction and not an approximation — the Appendix C draft's "$\approx$"
in its equation (C.13) understates it, and with equality the correspondence to
@eq-rho-ctt is algebraic rather than asymptotic. The simulation, by contrast, obtains
$\sigma^2_\theta$ by **construction**: it draws $\theta$ from a distribution with variance
one and reads reliability off @eq-wbar below.

The two operations may target the same population variance under the measurement-error
model, but they do not coincide automatically in expectation. The subtraction estimator
inherits sampling and plug-in error, whereas the simulation value is fixed by design.
Reading a realized $\bar w$ off a sample of twenty people when $\sigma^2_\theta$ was set by
fiat therefore mixes a design parameter with an estimator, and the mixture is what makes
negative reliability impossible in simulation and commonplace in data
(@sec-ch08-negative). The distinction is logged as `C-004` and `C-005`; @sec-ch09 shows
where it bites in the design table.

With $\sigma^2_\theta$ known, the population reliability is

$$
\bar w = \frac{\sigma^2_\theta}{\sigma^2_\theta + \mathrm{MSEM}}
       \;\;\xrightarrow{\;\sigma^2_\theta = 1\;}\;\; \frac{1}{1 + \mathrm{MSEM}} ,
$$ {#eq-wbar}

which is the form the manuscript uses and the form @sec-ch09 inverts.

## Separation and strata {#sec-ch08-separation}

Reliability on the $[0,1]$ scale compresses badly at the top: the difference between
$0.90$ and $0.95$ is not perceptible as a number but is a halving of error variance.
Wright and Masters give two rescalings that do not compress.

::: {#prp-separation}
**Separation and strata.** *Restated from Wright and Masters (1982, § 5.5)*

The person separation index is the ratio of true spread to measurement error in the same
units,

$$
S = \frac{\sigma_\theta}{\mathrm{RMSEM}},
\qquad\text{equivalently}\qquad
\bar w = \frac{S^2}{1+S^2},
$$ {#eq-separation}

and the number of statistically distinct person strata is

$$
H = \frac{4S+1}{3} .
$$ {#eq-strata}

Wright and Masters write these as $G_P = SA_P/SE_P$ and $H_P = (4G_P+1)/3$.
:::

@eq-strata is not arbitrary and its constant is not a fitted one. It follows from a stated
convention: two person locations count as distinct when they lie **three measurement
errors apart**, which places stratum centres at $3\,\mathrm{RMSEM}$ intervals across the
functional range of roughly $\pm 2\sigma_\theta$. That convention is the assumption to
argue with, not the algebra. A reader who prefers a two-error criterion gets a different
constant and a different count of strata from identical data.

$S$ has the advantage that it is unbounded and roughly linear in what one cares about:
$\bar w = 0.5$ is $S = 1$, meaning true spread and error are equal, and $\bar w = 0.9$ is
$S = 3$.

::: {.callout-note appearance="simple"}
::: {.ms}
**A defect in the manuscript's Table 2.** The table reports RMSEM, $S$, $H$, and test
information for five target reliabilities. Three of the four columns are computed from
the exact target and reproduce it to the displayed precision. The separation column is
not: it is computed from the *rounded* RMSEM column, and half-up rounding of
$1/\mathrm{RMSEM}_{\text{printed}}$ reproduces the printed $1.0/1.3/1.4/2.0/3.3$ exactly,
where the exact values are $1.0/1.2/1.5/2.0/3.0$. The printed $H$ column, computed from
the *exact* $S$, is internally inconsistent with the printed $S$ beside it. The
manuscript's own prose settles which is intended: at $\bar w = 0.9$ it says the
between-person SD is about three times RMSEM, which is $S = 3.0$, not $3.3$. Logged as
`C-001`, severity minor; the fix is to regenerate the column from the target reliability.
:::
:::

## Three functionals, and the order they come in {#sec-ch08-functionals}

Everything above averaged over persons. At the population level the same quantities can
be assembled in three different orders, and the literature has settled on all three
without much noticing.

Write $\mathcal{J}(\theta)$ for the test information at $\theta$ and let $G$ have variance
$\sigma^2_\theta$. Then:

$$
\bar w = \frac{\sigma^2_\theta}{\sigma^2_\theta + \operatorname{E}_G[\mathcal{J}^{-1}]},
\qquad
\rho_\Theta = \operatorname{E}_G\!\left[\frac{\sigma^2_\theta\mathcal{J}}
                                            {\sigma^2_\theta\mathcal{J} + 1}\right],
\qquad
\tilde\rho = \frac{\sigma^2_\theta\operatorname{E}_G[\mathcal{J}]}
                  {\sigma^2_\theta\operatorname{E}_G[\mathcal{J}] + 1} .
$$ {#eq-functionals}

$\bar w$ averages the *error variances* and then forms the ratio. $\rho_\Theta$ forms the
ratio pointwise and averages *that*. Andersson and Xin
[-@andersson_large_2018, eq. 13], following Cheng et al. [-@cheng_comparison_2012], state
the unit-variance form; the $\sigma^2_\theta$ form displayed here is its algebraic scale
adaptation. `mirt::marginal_rxx` does **not** implement this general-variance formula: mirt
1.46.1 evaluates $\operatorname{E}_G[\mathcal{J}/(\mathcal{J}+\sigma^2_\theta)]$, which
agrees with $\rho_\Theta$ only when $\sigma^2_\theta=1$. $\tilde\rho$ averages the
*information* and then forms the ratio; it is the average-information metric that the
reliability-targeting package calibrates against.

They are not equal, and they are ordered.

::: {#thm-jensen-sandwich}
**The three functionals are sandwiched.** *Outer inequality restated from the
reliability-targeting package (`IRTsimrel`, `vignettes/theory-reliability.Rmd`, Theorem
1); the placement of $\rho_\Theta$ between the two is derived here*

For any $G$ with $\sigma^2_\theta > 0$ and any information function with
$\mathcal{J} > 0$ and $\operatorname{E}_G[\mathcal{J}]$, $\operatorname{E}_G
[\mathcal{J}^{-1}]$ finite,

$$
\bar w \;\le\; \rho_\Theta \;\le\; \tilde\rho ,
$$ {#eq-sandwich}

with equality throughout if and only if $\mathcal{J}(\theta)$ is constant
$G$-almost surely.
:::

*Proof.* Both steps are Jensen's inequality on the same function seen from two sides.
For the left inequality, $x \mapsto \sigma^2_\theta/(\sigma^2_\theta + x)$ is strictly
convex on $x > 0$, so applying it to $x = \mathcal{J}^{-1}$ gives $\operatorname{E}_G
[\sigma^2_\theta/(\sigma^2_\theta + \mathcal{J}^{-1})] \ge \sigma^2_\theta/
(\sigma^2_\theta + \operatorname{E}_G[\mathcal{J}^{-1}])$, and the left side is
$\rho_\Theta$ after multiplying through by $\mathcal{J}$. For the right inequality,
$J \mapsto \sigma^2_\theta J/(\sigma^2_\theta J + 1)$ is strictly concave on $J > 0$, so
$\operatorname{E}_G[\,\cdot\,] \le \,\cdot\,(\operatorname{E}_G[\mathcal{J}])$, which is
$\rho_\Theta \le \tilde\rho$. Strictness of the convexity gives the equality condition in
both. ∎

The outer inequality $\bar w \le \tilde\rho$ is the theorem the targeting package proves
and relies on; what @thm-jensen-sandwich adds is that the coefficient most often reported
in practice sits strictly between the two, so the three are not merely different but
rankable, and any comparison across papers that does not name the functional has an
unsigned bias of known direction.

**How much does it matter?** Less than the theorem's tidiness suggests, and the honest
answer is worth more than the theorem. @fig-jensen evaluates all three by quadrature over
Rasch tests of 5 to 95 items. Under a standard normal $G$ the outer gap peaks near
$0.008$ around $\bar w = 0.6$ and falls below $0.003$ by $\bar w = 0.9$; the inner gap is
roughly half of that. Across four shapes of $G$ the largest gap anywhere in the grid is
about $0.021$. For design work at the precision anyone reports reliability to, the three
functionals are interchangeable; the reason to keep them distinct is that the *direction*
is guaranteed and the magnitude is not, and a target met on one metric is met with a
known sign of error on the others.

![The three marginal reliability functionals of @sec-ch08-functionals, over Rasch tests of 5 to 95 items with difficulties evenly spaced on $[-1.6,1.6]$. Left: the two functionals plotted as distances above $\bar w$, so the sandwich is visible; $\rho_\Theta$ lies strictly between the other two everywhere, and the inner gap is roughly half the outer one. Note the vertical scale. Right: the outer gap $\tilde\rho-\bar w$ under four latent distributions, each standardized to variance 1. Heavy tails widen the gap and bimodality narrows it relative to the normal, because the gap tracks how much information varies over the mass of $G$, not how far $G$ is from normal. Every gap here is below 0.03. Deterministic quadrature; no simulation. Generated by code/R/09-figures.R.](../figures/F-jensen.png){#fig-jensen}

The right panel carries a result that is easy to state backwards. The gap does not track
how far $G$ is from normal. It tracks how much $\mathcal{J}$ varies over the mass of
$G$ — which is what the equality condition in @thm-jensen-sandwich says. Heavy tails
widen it, because tail mass sits where information is scarce and $\mathcal{J}^{-1}$ is
enormous. Bimodality *narrows* it below the normal case, because two modes at
$\pm 1.2$ concentrate mass in the region the items cover. The flagship non-normal case of
this programme's own simulation is therefore the case where the choice of functional
matters least.

## What the conventions are worth {#sec-ch08-applied}

The thresholds $0.70$, $0.80$ and $0.90$ have no derivation. They are conventions, and
their provenance is thinner than their authority. Nunnally's textbook is the usual
citation and is regularly cited for a threshold he stated with qualifications about
research stage that the citing literature drops; Sijtsma [-@sijtsma_use_2009] makes the
point at length, and Cronbach's own retrospective on $\alpha$
[-@cronbach_my_2004] never names a numerical threshold at all — his argument is that
$\alpha$ should give way to a generalizability analysis, not that it should clear a bar.
Hussey et al.
[-@hussey_aberrant_2025] supply the empirical signature of a convention operating as a
gate rather than a description: across more than 67,000 published $\alpha$ values, an
excess at the $.70$ rule-of-thumb threshold that justifiable measurement practice does
not explain.

Against that, @sec-ch01-reliability gave the ranges actually observed — content-area
curriculum-based measures running from $.21$ to $.89$, PISA subscales as low as $.41$ in
some subgroups, and achievement batteries reliable at $.83$–$.92$ overall performing
materially worse for English language learners. The corpus census sharpens the point
into a distribution: across 889 Warehouse datasets, the share below $.80$ is 30
percent or 50 percent depending only on which of two defensible coefficients is
computed, and six estimators applied under identical rules span $.794$ to $.868$ in
their pooled values [@lee_reliability-distribution_2026] — the zoo of
@sec-ch08-zoo, operating at scale. Those numbers are the reason a book about
estimating person-specific traits cannot treat reliability as a hurdle already cleared.

Three closing statements, which the rest of the book uses:

1. A reliability coefficient without its name is uninterpretable, and with its name it
   is still uninterpretable without the population it was computed in.
2. On the latent scale the coefficient is a monotone transform of the separation index,
   and $S$ is the more honest display above $\bar w = 0.9$.
3. Reliability answers a question about the *ensemble*: what fraction of observed spread
   is real. It says nothing about whether any particular person's estimate is good, and
   @sec-ch17 shows that the estimator which maximizes it is not the estimator one wants
   for several other goals.

The framework here is the Rasch-centred reliability spine. @sec-ch25 does not replace it
with a parallel second development; it asks what changes under the 2PL when discrimination
heterogeneity makes information much less even and the functionals above can separate
more sharply.

## Sources and provenance {#sec-ch08-sources}

The corpus span and definition-dependent shares quoted in @sec-ch08-applied are the
Item Response Warehouse reliability study's [-@lee_reliability-distribution_2026],
read from its manuscript and taken up properly in @sec-ch28.

@eq-ctt and @eq-rho-ctt, the five equivalences, the over- and under-dispersion identities
@eq-overdisp and @eq-underdisp, the two reliability forms @eq-rel-subtract and
@eq-rel-add, and the measurement-design-effect framing are Adams
[-@adams_reliability_2005], who attributes the person-separation form to Wright and Stone
(1979) and the posterior-variance identity to Verhelst (2001) and to Mislevy et al.
[-@mislevy_estimating_1992]; neither Wright and Stone (1979) nor Verhelst (2001) is held,
and both are cited as secondary through Adams under R5.

@eq-alpha is Cronbach's own equation (2) [-@cronbach_coefficient_1951, p. 299], read at
that page: the symbols $n$, $V_i$ and $V_t$ and the statement that $V_t$ is the variance
of test scores and $V_i$ the variance of item scores are his. The lower-bound status
and the $\lambda_3 = \alpha$ identity are van der Ark [-@ark_standard_2025, pp. 1679 and
1686 § 3.7]; $\lambda_2$ is Guttman [-@guttman_basis_1945, § 13].

@eq-separation and @eq-strata, and the variance decomposition underlying
@sec-ch08-design, are Wright and Masters [-@wright_rating_1982, § 5.5], read directly;
the three-measurement-error convention behind @eq-strata is theirs and is stated here
because the constant in @eq-strata is otherwise unmotivated. `C-002` and `C-003` record
that the manuscript's attributions of $\bar w$ and of $H$ to that source are correct.

$\rho_\Theta$ in @eq-functionals adapts the unit-variance expression of Green et al.
[-@green_technical_1984] as given by Andersson and Xin
[-@andersson_large_2018, eq. 13] following Cheng et al. [-@cheng_comparison_2012]. The
outer inequality of @thm-jensen-sandwich is Theorem 1 of
the reliability-targeting package's theory vignette, restated; the inner placement and
the proof as organized here are derived, and neither is claimed as novel — both steps are
Jensen's inequality. @prp-mean-weight is derived here and corresponds to equations
(C.23)–(C.24) of the Appendix C draft, which gives the inequality but not, in its earlier
form, the working-approximation qualification; that qualification is logged as `C-006`.

Every implementation formula in @tbl-zoo was read from package source — `TAM::WLErel`
(TAM 4.3-25), `mirt::empirical_rxx` and `mirt::marginal_rxx` (mirt 1.46.1), and
`IRTsimrel::compute_rho_tilde` (0.2.0) — rather than from documentation or recollection.
Two notes follow from that reading. `mirt::marginal_rxx` evaluates $\mathcal{J}/
(\mathcal{J} + \sigma^2_\theta)$, which agrees with $\rho_\Theta$ in @eq-functionals only
at unit variance. Unit variance is not a Rasch identification requirement, so for a
nonunit latent variance this is an implementation mismatch with the standard scaled
coefficient unless a different estimand is explicitly intended. The numerical comparison
in @sec-ch08-zoo is a check run for this chapter, not a published result.

The empirical ranges in @sec-ch08-applied are carried forward from @sec-ch01-reliability
with their sources; nothing new is asserted about them here.
