# The Bayesian Hierarchical Rasch Model {#sec-ch10}

```{r}
#| label: setup-ch10
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

Parts II and III treated the item and person parameters as things to be estimated, and
the population distribution $G$ as either a device for eliminating incidental parameters
(@sec-ch05-mml) or a source of shrinkage (@sec-ch06-family). This part writes the model
that holds all of it together, and it does so for a specific reason: the argument of this
research programme is about the *shape* of $G$, and a hierarchical Bayesian formulation
is the framework in which that shape can be given a prior and learned. Everything in
Part V depends on the model stated here.

The chapter is deliberately concrete. It states the model the companion software actually
fits, read from its source, rather than a textbook idealization — and where the two differ,
it says so.

## Four levels, and what is actually at each one {#sec-ch10-levels}

The phrase "second stage," which the manuscript uses, is a source of avoidable confusion:
different literatures count stages differently. The level vocabulary of @sec-ch02-levels
is used instead.

**Level 1, the measurement model.** Conditional on $\theta_p$ and $\boldsymbol\beta$,
responses are independent Bernoulli variables with the Rasch probability of @eq-rasch:

$$
U_{pi} \mid \theta_p, \beta_i \;\stackrel{\text{ind}}{\sim}\;
\operatorname{Bernoulli}(\pi_{pi}), \qquad
\operatorname{logit}(\pi_{pi}) = \theta_p - \beta_i .
$$ {#eq-level1}

This is the only level that touches the data, and it carries the local-independence
assumption of @sec-ch03-lsi.

**Level 2, the persons.** $\theta_p \mid G \stackrel{\text{iid}}{\sim} G$. Under the
parametric specification $G = N(\mu_\theta, \sigma^2_\theta)$; under the semiparametric
one, $G$ is itself given a prior, which is @sec-ch16. Level 2 is where the entire book
turns, and stating it as $\theta_p \sim G$ rather than $\theta_p \sim N(\cdot)$ is the
whole of the generalization.

**Level 3, the items.** The item difficulties are given a common distribution as well.
This is where the implemented model and the drafted appendix part company, and
@sec-ch10-item-prior takes it up separately because the difference is not cosmetic.

**Level 4, the hyperpriors.** Priors on whatever remains free at level 2 and level 3.
@sec-ch10-hyperpriors reports what these actually are and what is known about the choice.

## The item prior is not the one usually written down {#sec-ch10-item-prior}

The Appendix D draft specifies level 3 as $\beta_i \mid \mu_\beta, \sigma^2_\beta
\stackrel{\text{iid}}{\sim} N(\mu_\beta, \sigma^2_\beta)$ with a flat prior on $\mu_\beta$
and an inverse-$\chi^2$ prior on $\sigma^2_\beta$. The fitted model does something else.
Read from the model code, the Rasch specification under item-centred identification is

$$
\tilde\beta_i \stackrel{\text{iid}}{\sim} N(0, \sigma^2_\beta),
\qquad
\beta_i = \tilde\beta_i - \frac{1}{I}\sum_{j=1}^{I}\tilde\beta_j ,
\qquad \sigma^2_\beta = 3 \text{ fixed},
$$ {#eq-item-centered}

with no $\mu_\beta$ anywhere and no hyperprior on $\sigma^2_\beta$. The centering is the
identification device of @sec-ch04-constraints imposed inside the sampler rather than
after it, and it makes the free item mean unnecessary — which is why the model does not
have one.

The prior on $\boldsymbol\beta$ that this induces is worth writing out, because it is not
what @eq-item-centered looks like it is.

::: {#prp-centered-item}
**The induced item prior is singular, and its marginal variance is not $\sigma^2_\beta$.**
*Derived here; no originality claim*

With $\tilde{\boldsymbol\beta} \sim N(\mathbf{0}, \sigma^2_\beta \mathbf{I}_I)$ and
$\boldsymbol\beta = (\mathbf{I}_I - I^{-1}\mathbf{1}\mathbf{1}^{\top})
\tilde{\boldsymbol\beta}$, the induced law of $\boldsymbol\beta$ is Gaussian with mean
zero, covariance $\sigma^2_\beta(\mathbf{I}_I - I^{-1}\mathbf{1}\mathbf{1}^{\top})$, and
rank $I-1$. In particular

$$
\operatorname{Var}(\beta_i) = \sigma^2_\beta\!\left(1 - \frac{1}{I}\right),
\qquad
\operatorname{Cov}(\beta_i, \beta_j) = -\frac{\sigma^2_\beta}{I},
\qquad
\operatorname{Cor}(\beta_i, \beta_j) = -\frac{1}{I-1}
$$ {#eq-centered-moments}

for $i \ne j$, and $\sum_i \beta_i = 0$ almost surely.
:::

*Proof.* The centering matrix $\mathbf{C} = \mathbf{I}_I - I^{-1}\mathbf{1}\mathbf{1}
^{\top}$ is symmetric and idempotent with rank $I-1$, so $\boldsymbol\beta = \mathbf{C}
\tilde{\boldsymbol\beta}$ is Gaussian with covariance $\sigma^2_\beta \mathbf{C}
\mathbf{C}^{\top} = \sigma^2_\beta\mathbf{C}$. Reading off the diagonal and off-diagonal
entries of $\mathbf{C}$ gives @eq-centered-moments, and $\mathbf{1}^{\top}\mathbf{C} =
\mathbf{0}$ gives the sum constraint. ∎

Three consequences follow, and the first two are practical.

**Setting $\sigma^2_\beta$ does not set the item variance.** At $I = 12$ and
$\sigma^2_\beta = 3$, each $\beta_i$ has prior variance $2.75$, not $3$. The discrepancy
is $O(I^{-1})$ and negligible on long tests, but it is a real bias on the short tests
that @sec-ch09-table2 shows this programme uses at low reliability, and anyone eliciting
$\sigma^2_\beta$ from a belief about item spread should divide accordingly.

**The difficulties are negatively correlated a priori.** The correlation $-1/(I-1)$ is
small but structural: it is the price of the constraint, and it says that the prior
carries information about item *differences* and none at all about their common level.
That is exactly right, because the common level is not identified (@prp-location), and it
is a clean illustration of the general point that a constraint is a choice of
representative from an orbit rather than an assumption about the world.

**The prior is improper in no direction and degenerate in one.** The induced law lives on
the $(I-1)$-dimensional sum-to-zero hyperplane. There is no flat prior anywhere in the
model, which matters for the propriety questions @sec-ch10-hyperpriors raises.

::: {.callout-important}
## The appendix and the implementation describe different models

The Appendix D draft's level 3 has a free item mean $\mu_\beta$ with a flat prior and a
hyperprior on $\sigma^2_\beta$. The fitted model has neither: the mean is removed by
centering and the variance is fixed at 3. The drafted specification is strictly more
general than the one that produced the results.

Neither is wrong as a model. What is wrong is describing one and fitting the other, and
the appendix should be brought into line with @eq-item-centered before it is published.
Logged as `C-013`.
:::

## The joint posterior {#sec-ch10-posterior}

With the levels in place the posterior factors in the only way it can.

::: {#prp-joint-posterior}
**Joint posterior factorization.** *Adapted from Fox (2010, ch. 2, § 2.1)*

$$
p(\boldsymbol\theta, \boldsymbol\beta, \boldsymbol\psi \mid \mathbf{u})
\;\propto\;
\underbrace{\textstyle\prod_{p}\prod_{i} \pi_{pi}^{u_{pi}}(1-\pi_{pi})^{1-u_{pi}}}
           _{\text{level 1}}
\;\cdot\;
\underbrace{\textstyle\prod_{p} g(\theta_p \mid \boldsymbol\psi_\theta)}_{\text{level 2}}
\;\cdot\;
\underbrace{p(\boldsymbol\beta \mid \boldsymbol\psi_\beta)}_{\text{level 3}}
\;\cdot\;
\underbrace{p(\boldsymbol\psi)}_{\text{level 4}} ,
$$ {#eq-joint-posterior}

where $\boldsymbol\psi = (\boldsymbol\psi_\theta, \boldsymbol\psi_\beta)$ collects the
hyperparameters and the level-4 term factors further under prior independence.
:::

*Proof.* Bayes's rule applied to the joint density, which factors as stated because each
level is conditionally independent of everything above it given its own parents. ∎


Two features of @eq-joint-posterior earn their keep later. The level-2 term is the *only*
place $G$ enters, so replacing a normal $G$ by a Dirichlet process mixture changes one
factor and leaves the rest of the model untouched — that modularity is why the
semiparametric extension of @sec-ch16 is an extension rather than a rewrite. And the
level-3 term does **not** factor over items under @eq-item-centered, because
@prp-centered-item makes the $\beta_i$ dependent; writing it as $\prod_i p(\beta_i)$,
as the Appendix D draft does, is correct for the drafted model and false for the fitted
one.

## Hyperpriors, and what is known about choosing them {#sec-ch10-hyperpriors}

The implemented defaults, read from the model code, are

$$
\mu_\theta \sim N(0, 3), \qquad
\sigma^2_\theta \sim \text{Inv-Gamma}(2.01, 1.01) .
$$ {#eq-hyperpriors}

The second is worth a moment. An inverse-gamma with shape $a$ and scale $b$ has mean
$b/(a-1)$ and finite variance only for $a > 2$. At $(2.01, 1.01)$ the mean is exactly
$1.0$ and the variance is $100$. Fixing Rasch discriminations at one defines the unit; it
does **not** fix the population variance after location anchoring. The value one is
therefore substantive prior centring, not an identification constraint. The default is
proper and heavy-tailed, but its suitability must be checked by prior-predictive or
sensitivity analysis rather than inferred from identification.

It is also not the specification the standard warning is about.

::: {.callout-note appearance="simple"}
::: {.ms}
**What the inverse-gamma pathology does and does not say here.** Gelman
[-@gelman_prior_2006, §§ 2.1–2.2] shows that the inverse-gamma$(\epsilon, \epsilon)$
family — the $\text{IG}(0.001, 0.001)$ that older software offered as "noninformative" —
has *no proper limiting posterior* as $\epsilon \to 0$, so inferences remain sensitive to
$\epsilon$ and it "cannot simply be comfortably set to a low value." Elsewhere in the
paper Gelman develops uniform and half-$t$ alternatives on the standard-deviation scale;
that recommendation is not attributed to §§ 2.1–2.2.

Two things follow for this model. The default @eq-hyperpriors is *not* of the
$(\epsilon,\epsilon)$ form and does not inherit the pathology. And the "groups" at level 2
are persons, of which there are hundreds or thousands, so the regime where the choice
dominates — few groups, group-level variance near zero — is not the regime this programme
works in. The warning is real and is aimed elsewhere; repeating it here without that
qualification would be citation by topic adjacency rather than by content (R1).
:::
:::

The prior on $\sigma^2_\theta$ does still matter in one place this book cares about. It
sets the shrinkage weight of @eq-shrinkage-weight, hence the reliability of
@sec-ch08-shrinkage, and under a semiparametric $G$ its role is taken over by the
base-measure and concentration hyperparameters of @sec-ch15, where the sensitivity is
larger and less well understood.

## A prior and a population distribution are still not the same thing {#sec-ch10-parallel}

@sec-ch02-frameworks laid out the parallel and @tbl-frameworks-ch10 repeats it, because
this is the chapter where the temptation to collapse the two is strongest.

```{r}
#| label: tbl-frameworks-ch10
#| tbl-cap: "The role of $G$ in the two frameworks, repeated from @sec-ch02-frameworks. Source: tables/T-frameworks.rds."
#| echo: false
knitr::kable(tbl("T-frameworks"))
```

The mathematics is genuinely shared. Marginal maximum likelihood integrates the person
likelihood against $g$ (@eq-mml); the hierarchical Bayesian model multiplies by the same
$g$ at level 2 of @eq-joint-posterior. MML then maximizes over $(\boldsymbol\beta,
\boldsymbol\psi_\theta)$ while the Bayesian model integrates over them under a
hyperprior, and MML is recoverable as the limit of flat hyperpriors with a posterior
concentrating at the maximizer.

What differs is the epistemic status and what follows from it. Under MML the normality of
$G$ is a structural claim about a population, assessable by fit and relaxable by
nonparametric MML. Under the Bayesian model it is a prior belief, updated by data and
relaxable by a nonparametric prior. Reviewer 2 was right that a distributional assumption
is not a specifically Bayesian sin; @sec-ch02-frameworks says so and this chapter does not
retreat from it.

The honest addition is that **the practical difference is smaller than the rhetorical
one**. Both frameworks shrink, both are sensitive to a misspecified $G$ in the same
direction, and @sec-ch12 shows that the shrinkage phenomenon is a property of the
hierarchical structure rather than of the inferential philosophy. What the Bayesian
formulation buys is not correctness. It is three specific things.

## Why the Bayesian formulation, specifically {#sec-ch10-why-bayes}

**Uncertainty propagates coherently from items to persons.** This is the argument
@sec-ch06-plugin set up and `C-012` recorded. Two-stage practice calibrates items, fixes
them at their estimates, and scores persons as though the item parameters were known;
@eq-sem is then a conditional standard error reported as though it were unconditional.
Patz and Junker [-@patz_straightforward_1999] state the alternative exactly: MCMC "treat[s]
item and subject parameters at the same time," which "allows us to incorporate standard
errors of item estimates into trait inferences, and vice versa." That is the plug-in gap
closed by construction rather than by an extra term, and it is the cleanest argument in
the list.

**Small samples.** Where $P$ is small the incidental-parameter difficulty of
@thm-neyman-scott is at its worst and the asymptotics of @eq-sem are least trustworthy.
Gilholm et al. [-@gilholm_bayesian_2021] fit hierarchical multidimensional IRT to small,
sparse samples for exactly this reason, and Fox [-@fox_redefining_2025] takes the more
radical route of integrating the person factor out analytically to obtain a model that
behaves in small samples. Neither is this programme's setting — the simulation uses
hundreds of persons — so the argument is real but is not what carries the case here, and
overstating it would be easy.

**The route to a nonparametric $G$.** This is the argument that actually matters. Under
MML, relaxing $G$ means nonparametric maximum likelihood, whose solution is a discrete
mixing distribution on finitely many support points — useful, and not a density. Under the
hierarchical model, relaxing $G$ means putting a prior on it, and the Dirichlet process
mixture of @sec-ch15 delivers a continuous density with a full posterior over its shape.
Since the question this programme asks is what the *distribution* of latent traits looks
like and how well its features are recovered, a framework that yields a posterior over
distributions is not a stylistic preference. It is the object of study.

## Computation, briefly {#sec-ch10-computation}

@eq-joint-posterior has no closed form and is sampled. Four approaches appear in this
literature, and the differences among them are not merely tactical.

Albert [-@albert_bayesian_1992] introduced data augmentation for the normal-ogive model:
latent continuous responses are imputed, conditional on which the item parameters have
conjugate normal full conditionals, so the whole sampler is Gibbs. It is elegant and it is
tied to the probit link. Patz and Junker [-@patz_straightforward_1999] gave the
general-purpose alternative — Metropolis-within-Gibbs, which does not require conjugacy
and therefore extends to models that have no augmentation trick, which is most of them.
Hamiltonian Monte Carlo as implemented in Stan is the modern default for the parametric
model [@burkner_bayesian_2021]. The semiparametric models of Part V are fitted in NIMBLE,
which is what makes the Dirichlet process machinery available at all.

The identification strategy interacts with the sampler, and @sec-ch04-constraints already
said why: centering inside the sampler (@eq-item-centered) and rescaling afterwards target
the same identified posterior for every location-invariant function, but they present the
sampler with different geometry and can differ in efficiency by an order of magnitude.
Details are in Appendix F; the point for here is that this is a computational choice with
no inferential content, and it should not be mistaken for a modelling one.

## Sources and provenance {#sec-ch10-sources}

The four-level structure follows the Appendix D draft §§ D.2.1–D.2.5. Fox
[-@fox_bayesian_2010, ch. 2, § 2.1, eqs. 2.1–2.2] directly gives the conditional-
independence factorization for person and item hierarchies; @prp-joint-posterior adapts it
to the four levels and centred item prior used here. Its inline proof records the algebra;
the `adapted` provenance records the literature source rather than claiming priority.

@eq-item-centered and @eq-hyperpriors were read from the companion package's model source
(`R/model_spec.R`, the item-centred Rasch specification and the default constants), not
from its documentation or its vignette, and the discrepancy with the Appendix D draft that
@sec-ch10-item-prior reports was found that way. @prp-centered-item is elementary linear
algebra, derived here, and was checked numerically before being written.

The inverse-gamma discussion is Gelman [-@gelman_prior_2006, §§ 2.1–2.2], read directly;
the qualification that its regime is small-$J$ and that @eq-hyperpriors is not of the
$(\epsilon,\epsilon)$ form is this book's, and is stated because the citation is otherwise
routinely made without it. The moments of $\text{Inv-Gamma}(2.01, 1.01)$ are computed from
the standard formulas.

The uncertainty-propagation argument quotes Patz and Junker [-@patz_straightforward_1999];
the small-sample arguments are Gilholm et al. [-@gilholm_bayesian_2021] and Fox
[-@fox_redefining_2025], read at the level of their stated aims rather than their
derivations, which is all the claim here requires. Albert [-@albert_bayesian_1992] is
cited for data augmentation under the normal ogive and Bürkner
[-@burkner_bayesian_2021] for the Stan route.

The parallel-structure argument of @sec-ch10-parallel is @sec-ch02-frameworks restated
with the same sources; the added claim — that the practical difference is smaller than the
rhetorical one — is this book's and is supported in @sec-ch12 rather than here.
