# Three Goals, Three Losses {#sec-ch17}

```{r}
#| label: setup-ch17
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
cx17 <- tbl("T-three-goals-counterexample")
```

Reviewer 2's sharpest objection to the manuscript was that its estimators are "reused with
limited adaptation." The answer is that the adaptation is a matter of **which loss
function an assessment context implies**, and that argument has to be made rather than
asserted. This chapter makes it.

Everything before this point in the book has been about *estimating* — how to get
$\hat\theta_p$, how precise it is, what happens when $G$ is misspecified. Part VI asks a
prior question: **what are you estimating for?** Three answers recur in assessment, they
imply three different losses, and @thm-incompatible shows by existence that their Bayes
actions need not be representable by one set of person estimates.

## Bayes estimators are loss minimizers {#sec-ch17-bayes}

The general statement is one line. Given a posterior $p(\boldsymbol\theta \mid \mathbf{u})$
and a loss $L(\mathbf{a}, \boldsymbol\theta)$ measuring the cost of taking action
$\mathbf{a}$ when the truth is $\boldsymbol\theta$, the **Bayes action** minimizes the
posterior expected loss,

$$
\mathbf{a}^{\star} = \arg\min_{\mathbf{a}}
\operatorname{E}\!\left[L(\mathbf{a}, \boldsymbol\theta) \mid \mathbf{u}\right] .
$$ {#eq-bayes-action}

The posterior is common to all three goals. **The loss is not**, and @eq-bayes-action is
the whole reason a single posterior yields three different answers. A reader who takes
"the Bayesian estimate" to be one object has skipped the loss.

Two consequences frame the chapter. Changing the loss changes the estimator even though
nothing about the model or the data has changed — so an estimator cannot be defended
without naming the decision it serves. And an estimator that is optimal for one loss is
generally *not* merely worse for another but wrong in a specific, predictable direction,
which is what makes the trade-off analysable rather than a matter of taste.

## Goal 1 — the individual {#sec-ch17-individual}

The first goal is the familiar one: report a number for each person that is close to that
person's ability. The loss is weighted squared error, summed over persons.

::: {#prp-pm-bayes}
**The posterior mean is Bayes under squared-error loss.** *Restated from Shen and Louis
(1998, §§ 2.2, 2.4, pp. 457--458)*

With $\mathrm{WSEL}(\mathbf{a}, \boldsymbol\theta) = \sum_p w_p (a_p - \theta_p)^2$ and
$\eta_p = \operatorname{E}[\theta_p \mid \mathbf{u}]$, $v_p =
\operatorname{Var}[\theta_p \mid \mathbf{u}]$,

$$
\operatorname{E}[\mathrm{WSEL} \mid \mathbf{u}]
 = \sum_p w_p (a_p - \eta_p)^2 + \sum_p w_p v_p ,
$$ {#eq-wsel}

so the minimizer is $a_p = \eta_p$ for every $p$ and every weighting, and the residual
$\sum_p w_pv_p$ is the irreducible posterior risk. Shen and Louis use a person-indexed
lambda for this quantity; this book uses the registered $v_p$ so that $\lambda_i$ remains
reserved for 2PL item discrimination (TD-3).
:::

@eq-wsel is worth reading twice, because the decomposition does the argument's work. The
first term is the only part the analyst controls; the second is what the data leave
uncertain. **Nothing an estimator does can reduce the second term** — which is why an
estimator that improves some other criterion must be paying for it in the first term, and
why the size of that payment is a computable quantity rather than a matter of opinion.

This is the EAP of @thm-eap-shrinkage arriving from the decision side, and it is the
estimator the manuscript uses. @sec-ch11 derived it; this chapter says what it is *for*.

## Goal 2 — the ranks {#sec-ch17-ranks}

The second goal reports an ordering or percentile rank: how a class sorts and how uncertain
each position is. The parameter of interest is the rank vector
$R_p = \sum_q \mathbb{1}\{\theta_q \le \theta_p\}$, and the natural loss is squared error
*on ranks*, whose Bayes action is $\operatorname{E}[R_p \mid \mathbf{u}]$ — optimally
discretized back to a permutation, which is what Shen and Louis
[-@shen_triple-goal_1998, eq. 1] call $\hat R_p$.

::: {#prp-rank-of-mean}
**The rank of the mean is not the mean of the rank.** *Restated from Shen and Louis (1998,
§§ 2.1, 3) and Goldstein and Spiegelhalter (1996)*

$\operatorname{rank}(\operatorname{E}[\theta_p \mid \mathbf{u}])$ and
$\operatorname{E}[\operatorname{rank}(\theta_p) \mid \mathbf{u}]$ are different
functionals, and the second is the Bayes action under rank loss. They differ in **value in
general**, because expected ranks are pulled toward the middle of the range; they can agree
in deterministic, tied, symmetric, or limiting cases. Whether they differ in **induced
ordering** depends on the model.
:::

The value/ordering distinction is the one @sec-ch12-ranks establishes and this chapter must
not blur, because the book got it wrong once. Under the common-form Rasch model the
posteriors are stochastically ordered in the total score, and Shen and Louis
[-@shen_triple-goal_1998, Theorem 2] then give
$R^{*} = R^{\dagger} = \tilde R = \hat R$ up to reversal: every rank procedure induces the
same ordering. What still differs is what gets *reported* — the expected rank of the top
scorer is not $P$, and an interval around it is not degenerate.

So Goal 2 in this book's setting is **not** "the ordering is wrong and needs repair." It
is: the ordering is right, the rank *values* and their uncertainty are not what ranking
point estimates suggests, and the regimes where even the ordering fails — different test
forms, missing designs, integrated item uncertainty — must be named rather than assumed
away. @sec-ch20 is where that is developed.

The clean ordering result has a fixed-known-2PL analogue, but not an unrestricted 2PL
analogue. As @prp-nosuff establishes, conditional on one common set of known
discriminations, the 2PL likelihood is a
one-parameter exponential family in the weighted score
$T_{\boldsymbol\lambda}(\mathbf{u}_p)=\sum_i\lambda_i u_{pi}$; hence response patterns are
monotone-likelihood-ratio ordered by that score, and their posteriors inherit the same
stochastic order under a common prior. If discriminations or other item parameters are
estimated and their uncertainty is integrated, the ordering statistic itself is no longer
fixed. Different forms, missing items, and person-dependent item-parameter posteriors then
need a separate ordering argument; this chapter makes no guarantee for those cases.

## Goal 3 — the distribution {#sec-ch17-distribution}

The third goal reports a distribution: what fraction of the population is below a cut
score, how spread out the class is, whether the trait is bimodal. Shen and Louis
[-@shen_triple-goal_1998, § 2] take the target to be the **empirical distribution of the
realized parameters**,

$$
G_K(t) = \frac{1}{K}\sum_{k} \mathbb{1}\{\theta_k \le t\},
$$ {#eq-edf-target}

and the loss to be integrated squared error, $\mathrm{ISEL}(A, G_K) = \int \{A(t) -
G_K(t)\}^2\,dt$. Their Theorem 1 gives the Bayes action: the minimizer of
$\operatorname{E}[\mathrm{ISEL} \mid \mathbf{Y}]$ is the posterior expectation of $G_K$,
and if the action is constrained to be a discrete distribution with at most $K$ mass
points, the minimizer $\hat G_K$ puts mass $1/K$ at points they characterize explicitly.

::: {.callout-important}
## $G_N$ is the realized EDF, not $G$ — R1's question, answered

Reviewer 1 wrote, at manuscript p. 16 line 32: "I became a bit lost in trying to understand
whether the quantity $G_N(t)$ is a rank estimate or a latent trait estimate. It appears to
be a mean probability of some type but then the authors describe it as a latent trait
estimate."

The confusion is the manuscript's to fix, not the reviewer's to overcome. $G_N$ is neither
a rank estimate nor a latent trait estimate. It is the **empirical distribution function of
the $N$ realized abilities in this sample** — @eq-edf-target with $K = N$ — so it is a
random object even given $G$, and estimating it is a third decision problem alongside
estimating the $\theta_p$ and estimating their ranks.

The distinction that makes it a *separate* estimand matters and is easy to lose:

- $G$ is the population distribution, a fixed unknown (frequentist) or a random measure
  with a prior (Bayesian). @sec-ch16 says a finite test identifies only $I+1$ of its
  functionals.
- $G_N$ is the EDF of the particular $N$ people who took the test. It converges to $G$ as
  $N$ grows, but for the $N$ at hand it is a different target, and it is the one a
  cut-score report is actually about.

The manuscript targets $G_N$ and the revision should say so in a sentence.
:::

## Names used in empirical comparisons {#sec-ch17-criteria}

The three goals acquire short labels in simulation tables, and those labels should not be
left implicit. If $A_p$ estimates the true rank $R_p$ among $P$ units, the mean squared error
loss of ranks is

$$
\mathrm{MSELR}=\frac{1}{P}\sum_{p=1}^{P}(A_p-R_p)^2.
$$

The mean squared error loss of percentile ranks rescales both ranks by $P$,

$$
\mathrm{MSELP}=\frac{1}{P}\sum_{p=1}^{P}
\left(\frac{A_p}{P}-\frac{R_p}{P}\right)^2
=\frac{1}{P^2}\mathrm{MSELR}.
$$

MSELP is therefore comparable across settings with different numbers of units in a way the
raw-rank scale is not [@lee_improving_2025, eqs. 6--7]. For an estimated EDF $\widehat G_P$
and target EDF $G_P$, the Kolmogorov--Smirnov distance is

$$
\mathrm{KS}(\widehat G_P,G_P)=\sup_t|\widehat G_P(t)-G_P(t)|.
$$

KS is a supremum diagnostic, whereas ISEL integrates squared discrepancy. Neither rank loss
nor an EDF distance is a substitute for individual RMSE; the names simply make the empirical
counterparts of the three losses explicit.

## The three goals can be incompatible {#sec-ch17-incompatible}

The paper's thesis is that the choice among posterior summaries matters. That is currently
asserted. The defensible theorem is an existence statement, not a universal boundary on all
overlapping posteriors.

::: {#thm-incompatible}
**A single trait vector need not minimize all three reporting losses.** *Adapted from Shen
and Louis (1998, §§ 2.1--2.4); existence proof here*

There exist independent, non-degenerate, overlapping posteriors for which no vector of trait
values can both be WSEL-optimal person by person and carry the ISEL-optimal $K$-point
distribution. It follows that no such vector can simultaneously attain those two optima and
the rank-loss-optimal ordering.
:::

*Proof.* Take $K = 2$ with independent posteriors $N(-0.6, 0.5)$ and $N(0.6, 0.2)$, where
the second normal argument is a variance. The unique WSEL action is the vector of posterior
means. Shen and Louis's Theorem 1 makes the ISEL-optimal equal-mass atoms the .25 and .75
quantiles of

$$
\bar G_2(t)=\tfrac12\Pr(\theta_1\le t\mid Y_1)
             +\tfrac12\Pr(\theta_2\le t\mid Y_2),
$$

and those atoms differ from the posterior means. Therefore no one trait vector attains both
optima, hence none attains all three. ∎

```{r}
#| label: tbl-three-goals-counterexample
#| tbl-cap: "The chapter-17 existence example. ISEL is evaluated against the posterior mean EDF. Source: tables/T-three-goals-counterexample.rds."
#| echo: false
z <- cx17
for (j in 2:ncol(z)) z[[j]] <- sprintf("%.6f", z[[j]])
knitr::kable(z)
```

The third row is included to prevent the original error from returning: posterior
expectations of the order statistics optimize a different ordered-location criterion, not
Shen and Louis's ISEL. In this example the ISEL action is more spread than the posterior
means, but @thm-incompatible does **not** make that direction universal. Some non-degenerate,
overlapping posterior configurations make the midpoint quantiles coincide with the means;
the theorem claims existence of conflict, not a degenerate-only equality boundary. How
often, and how expensively, the conflict is realized at practical designs is an
empirical question the theorem cannot answer — and now has a grid-scoped answer. The
companion simulation's pooled H7 contrast prices the trade in the theorem's direction;
the separate descriptive point census finds that direction at 239 of 240
condition-by-prior points, with one boundary-near KS exception (@sec-ch27). The
existence result stays an existence result; what changed is that the conflict has been
priced without turning the grid into a universal theorem.

Shen and Louis [-@shen_triple-goal_1998, § 3, p. 459] add the fact that determines the shape of
Part VI: **the ranks of the constrained-Bayes estimates are always identical to the ranks
of the posterior means**, because CB is a positive affine transformation of them. So a
spread repair cannot by itself be a rank repair. Their triple-goal construction is a
three-stage compromise targeted to the distributional and rank goals — estimate the EDF, minimize a rank loss,
then assign the $k$-th ranked unit the $\hat R_k$-th mass point of $\hat G_K$ — rather than
assuming all three unconstrained optima coincide. Its coordinate squared-error placement
interpretation needs the rank-agreement condition stated in @sec-ch19-properties.
@sec-ch18 and @sec-ch19 are those two routes.

## Which decision implies which loss {#sec-ch17-decisions}

The formalism is only useful if a practitioner can tell which case they are in.

**Squared-error loss on individuals** is implied when a number attaches to a person and is
acted on person by person: a placement decision, a diagnosis, an individual growth report.
The cost of error is borne by that person and does not depend on anyone else's estimate.

**Rank loss** is implied when the reported action is the ordering or the ranks themselves
and the trait scale is incidental: a published ordering, percentile-rank report, or league
table. Goldstein and Spiegelhalter [-@goldstein_league_1996] show why the uncertainty must
travel with the ranking; @sec-ch20 develops the decision problem.

Top-$k$ allocation and classification below a substantive cut are related but different
actions. A fixed-cardinality selection loss must price which units are included, while a
threshold-classification loss must price false negatives against false positives. Neither is
implied merely by writing down squared error on rank values.

**Integrated squared error on the EDF** is implied when the report is about the
*collection*: the proportion below a cut score, the spread of a class, whether a
distribution is bimodal, an estimated variance component. @sec-ch01 opened with this case
because it is the one where using the wrong estimator produces an answer that is
confidently and systematically wrong rather than merely noisy.

@tbl-goals collects the three.

```{r}
#| label: tbl-goals
#| tbl-cap: "Three goals, three losses, three Bayes actions. Source: tables/T-goals.rds."
#| echo: false
knitr::kable(tbl("T-goals"))
```

The table deliberately keeps classification out of the rank row. @sec-ch20 shows that a
rank estimator can be Bayes-optimal for rank loss and still be the wrong classifier.

A single assessment programme often needs all three at once, from one calibration. That is
not a contradiction; it is the reason @thm-incompatible has to be stated. **The estimates
should differ because the questions differ**, and reporting one vector of numbers for all
three purposes is the practice this book's Part VI exists to interrogate.

## Sources and provenance {#sec-ch17-sources}

@eq-bayes-action is standard Bayesian decision theory and is restated without attribution
to a particular text.

@prp-pm-bayes is Shen and Louis [-@shen_triple-goal_1998, § 2], read directly; the
decomposition @eq-wsel including the irreducible $\sum_p w_pv_p$ term is theirs.
@eq-edf-target, the ISEL loss, and the Theorem 1 characterization of its minimizer are
from their § 2.3, pp. 457--458, read directly. The rank-loss action $\hat R_p$ is their
equation (1) in § 2.1, and the CB rank-invariance fact quoted in
@sec-ch17-incompatible is their § 3, p. 459.

@prp-rank-of-mean is restated from Shen and Louis [-@shen_triple-goal_1998, §§ 2.1, 3] together
with Goldstein and Spiegelhalter [-@goldstein_league_1996]. **Its statement is deliberately
split into value and ordering**, because an earlier draft of this book asserted that
differential shrinkage reorders posterior means in the common-form Rasch model, and that
is false: Shen and Louis's Theorem 2 and Laird and Louis's [-@laird_empirical_1989,
Appendix A] Theorem A give coincidence of the rank procedures under stochastic ordering,
which @sec-ch12-ranks shows the Rasch total-score posterior satisfies. The correction is
recorded in the project's step logs.

@thm-incompatible is adapted. Paddock et al. [-@paddock_flexible_2006, abstract and § 2.1]
state the general motivation that no single estimate is optimal for all three goals, and
Shen and Louis supply each Bayes action. Neither source states the false universal condition
that an earlier version of this chapter attached to that motivation. The narrower existence
result and its normal example are this book's; the midpoint-quantile identities and direct
ISEL comparison are checked in `code/R/15-derivation-checks.R` and frozen in
`tables/T-three-goals-counterexample.rds`.

R1's question about $G_N$ is quoted verbatim from the reviewer comments at manuscript
p. 16 line 32. The reading of $G_N$ as the realized EDF follows from
@eq-edf-target; that the manuscript should say so explicitly is this book's recommendation
and is logged for the revision.
