# Relation to Lee et al. (2025) {#sec-ch21}

```{r}
#| label: setup-ch21
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

The Associate Editor's first demand was a clear delineation from Lee et al.
[-@lee_improving_2025], and Reviewer 2's fifth comment put it more sharply: the machinery
looks reused with limited adaptation. This chapter answers both, and it answers them by
stating the overlap first and in full. Understating what transfers is what invites the
charge; the delineation has to survive a reader who knows the earlier paper well.

Everything below is read from the source itself. The delineation table's cells, the
numbers, and the characterization of what that paper found are taken from the article at
the locators recorded in `manifest/part-vii-source-receipts.csv`, not from the
manuscript's summary of it, and not from the reviewer's. Where this chapter reports a work
that Lee et al. themselves cite — Rubin's model, Antonelli et al.'s diffuse prior,
Dorazio's induced prior on $K$ — it names the work without citing it, because what has
been read is their use of it and not the original. @sec-ch12 and @sec-ch15 cite those
sources on their own footing. That discipline changed the
chapter: one cell the plan for this book had written in advance turned out to be false,
and it was the cell the contribution argument had been resting on.

## What Lee et al. (2025) did {#sec-ch21-summary}

The setting is a multisite trial: (nearly) identical randomized experiments run at $J$
sites, analysed as what they call a Rubin model for parallel randomized experiments —
equivalently a random-effects meta-analysis. The model consumes two summary statistics per
site — the site's estimated
effect $\hat\tau_j$ and its squared standard error $\widehat{se}^2_j$, both computed by
maximum likelihood from site $j$'s own data — and treats the second as known:

$$\hat\tau_j \mid \tau_j, \widehat{se}^2_j \sim N(\tau_j, \widehat{se}^2_j),
\qquad \tau_j \mid G \stackrel{\text{i.i.d.}}{\sim} G .$$

Their target is the *finite-population* collection $\{\tau_j\}_{j=1}^J$ — the effects at
the sites actually in the trial — rather than the superpopulation $G$ from which those
sites were drawn. They make that choice deliberately, on the grounds that experimental
sites are often not representative of any wider population and that finite-population
inference is statistically less demanding.

Against that model they cross two strategies. The first relaxes the shape of $G$ with a
Dirichlet process prior in two arms: **DP-diffuse**, a deliberately vague $\alpha$ prior
they take from Antonelli et al. (2016), and **DP-inform**, their own construction, which
encodes a prior belief about the number of clusters $K$ as a $\chi^2(u)$ distribution and
then chooses the Gamma$(a,b)$ prior on $\alpha$ minimizing the Kullback–Leibler
divergence from the induced $\Pr(K \mid J, a, b)$ they take from Dorazio (2009). The second strategy replaces the posterior mean with a
summary targeted at a particular loss: constrained Bayes [@ghosh_constrained_1992] or
triple-goal [@shen_triple-goal_1998]. Three priors by three summaries gives nine
estimators, evaluated over 1,500 simulated conditions with 100 replications each and
summarized by meta-model regressions.

Their three inferential goals — estimating the individual effects, ranking the units, and
estimating the empirical distribution function — are **Shen and Louis's**, and Lee et al.
attribute them there explicitly. @sec-ch17 develops the same three, from the same source.
This book's manuscript should make that attribution as visible as Lee et al. do.

### Their findings, stated as they stand {#sec-ch21-findings}

Four results matter for what follows, and three of them constrain what this work may claim.

**Reliability governs everything.** The paper defines *informativeness*

$$I = \frac{\sigma^2}{\sigma^2 + \exp\!\left(\frac{1}{J}\sum_{j=1}^{J}
\ln \widehat{se}^2_j\right)},$$

which is the average reliability of the maximum-likelihood site estimates, and concludes
that it was the most influential factor for all three goals. They give prospective design
advice on that basis: when budget is constrained, increase the average number of subjects
per site.

**The prior choice reverses with reliability.** Semiparametric priors repay their cost
only when the data are informative. Below roughly $I = .20$ the Gaussian model paired with
a suitable summary performed on par with or better than the DP models **even when the true
$G$ was not Gaussian**. The crossover depends on the number of sites: at $J = 300$ the DP
models beat the Gaussian on integrated squared-error loss above about $I = .20$; at
$J = 75$ or $100$ they need about $I = .40$; at $J = 25$ they never significantly beat it,
even at the highest informativeness the design reached.

**The summary choice is goal-specific and its ordering is stable.** The posterior mean is
consistently best for root-mean-squared error and rarely best for the EDF; constrained
Bayes and triple-goal shrink less and represent the EDF better. In their second case study
the summary choice mattered more than the model choice: PM improved RMSE by 16–42 % over
CB or GR, while Gaussian and DP-inform differed by 1 % under PM.

**Ranking barely moves.** Mean-squared error loss of the percentiles was largely
unaffected by the number of sites, by the coefficient of variation of site sizes, by the
model, and by the summary method; it responded only to the average site size and to
$\sigma$. Their explanation is that different models and summaries seldom alter the
*ordering* of the estimates.

That last finding is qualitatively consistent with something this book had to correct in
itself. @sec-ch12 originally asserted that differential shrinkage reorders posterior
means; @sec-ch12 now shows it cannot in the common-form Rasch model, because the posterior
depends on the data only through the total score and the family is ordered in it. Lee et
al. report empirical near-invariance of ordering in a different hierarchical model. The
conclusions point in the same direction, but the mechanisms and evidential status differ:
their MSELP result does not corroborate the Rasch theorem.

## The delineation {#sec-ch21-delineation}

```{r}
#| label: tbl-lee-delineation
#| tbl-cap: "This work against Lee et al. (2025), read from the source. Source: tables/T-lee-delineation.rds."
#| echo: false
knitr::kable(tbl("T-lee-delineation"))
```

::: {.callout-important}
## A correction to this book's own plan

The blueprint for this chapter asserted that reliability is "not a design factor" in Lee
et al. **That is false**, and reading the paper is what showed it. Informativeness $I$ *is*
an average reliability, it is the paper's single most influential factor, and they give
design advice for raising it. The mechanism is the same one this book uses: more
observations per unit lowers the within-unit error and raises reliability, exactly as more
items lower the measurement error and raise $\bar w$.

The contribution argument had been resting on that cell. Reading the paper removes that
argument rather than replacing it with a claim about which reliability range belongs to
which setting. The defensible distinction is structural: this work uses the full
item-response likelihood, parameter-dependent information, and an identified latent
scale. Whether Lee et al.'s empirical pattern transfers as assessment reliability changes
is not established here; @sec-ch21-not-claimed assigns that empirical question.
:::

## What transfers {#sec-ch21-transfers}

```{r}
#| label: tbl-lee-transfer
#| tbl-cap: "Provenance of each element, and its status in this work. Source: tables/T-lee-transfer.rds."
#| echo: false
knitr::kable(tbl("T-lee-transfer"))
```

Four things transfer with no adaptation worth claiming credit for. The three-goal
framework is Shen and Louis's. Constrained Bayes is Ghosh's. The triple-goal estimator is
Shen and Louis's. The $K$-matched elicitation of $\alpha$ is Lee et al.'s own, and it is
used here as they built it: encode a belief about $K$, match a Gamma$(a,b)$ prior
to it, fit. A fifth — putting a Dirichlet process on the distribution of unit-level
parameters — is older than either paper and, in the item-response setting specifically,
has been proposed several times for different reasons (@sec-ch16).

Saying this plainly costs nothing. The question a delineation has to answer is not whether
the tools are shared but whether the problem is.

## What does not transfer {#sec-ch21-differences}

Three differences, each methodological rather than a re-labelling.

### The error variance is a function of the parameter {#sec-ch21-heteroscedastic}

Lee et al.'s first stage is a Gaussian sampling approximation motivated by within-site
large-sample theory. It plugs in $\widehat{se}^2_j$ from site $j$ and then treats that
estimate as known and fixed; doing so does not make the stage exact. Their model assumes
the site's precision is unrelated to the site's effect, and they name the assumption as a
key limitation of the study, suggesting three ways to relax it: fitting with and without
restrictions on the bivariate distribution of $\tau_j$ and $\widehat{se}^2_j$, treating
$\widehat{se}^2_j$ as a covariate in $G$, or applying a variance-stabilizing transform.

In an item-response model there is nothing to relax. The precision of a person's estimate
is the Fisher information $\mathcal{J}(\theta_p)$, which depends on $\theta_p$ — a person
near the difficulty of most items is measured precisely and a person at the extremes is
not (@sec-ch07). Shrinkage is therefore heteroscedastic by construction, and the normal
working model of @sec-ch11 is an approximation here where it was not needed there.

What this does *not* imply is worth stating, because this book got it wrong once. Varying
precision does not by itself reorder the posterior means: in the common-form Rasch model
the ordering is fixed by the total score (@sec-ch12). The consequence of heteroscedastic
shrinkage is in the *values and their uncertainty*, and in the regimes — different test
forms, incomplete designs, integrated item uncertainty — where the ordering does become
vulnerable (@sec-ch20).

### The latent scale has to be identified before $G$ can be read {#sec-ch21-identification}

A treatment effect is measured on the outcome's own scale. Lee et al. report effects in
effect-size units, and convert log odds ratios to Cohen's $d$ where the outcome is binary;
the scale is given by the data and no identification question arises.

A latent trait has no scale of its own. The location must be constrained, and under the
2PL the scale as well (@sec-ch04). The recovered $G$ is therefore a different object with
a different warrant, and before any comparison of priors can be interpreted one has to say
what a finite test identifies about $G$ at all. San Martín et al. [-@sanmartin_bayesian_2011]
answer it under their anchored Rasch conditions: the observable law is generated by
$I + 1$ displayed integral evaluations of $G$ (@sec-ch16). Those evaluations satisfy
algebraic identities and should not be read as $I+1$ unconstrained independent dimensions,
or as a proof that no other equivalent generating set can be written. Nothing in the
multisite setting raises this question.

This difference also has no analogue in the other direction. It is not that the multisite
problem is easier; it is that the question does not arise there, so the semiparametric
identification literature this book spends a chapter on has no place in that paper.

### Reliability is shared; its design mapping differs {#sec-ch21-reliability}

This is the contact point, and it needs care because the naive version of the claim is
false (@sec-ch21-delineation).

Both settings have a design lever that raises reliability, and both papers know it. In Lee
et al.'s simulations informativeness ranges from $.01$ to $.71$ with a mean of $.25$; their
applied example, a conditional cash transfer experiment in Bogotá with $J = 38$ sites,
sits at $I = .04$ and $.06$. According to Lee et al.'s comparison with Weiss et al.
(2017), the past multisite trials used to calibrate the design contain no case comparable
to their most informative simulated condition. That is a secondary attribution through
Lee et al., not an independently checked claim about Weiss et al.

The design mapping is nevertheless different. Lee's $I$ depends on average site size and
on $\sigma^2$, the true cross-site variance of effects; the latter is not a design choice.
Assessment reliability $\bar w$ depends on test information and on the target trait
population. Adding items can move it, but $I$ and $\bar w$ are not a common numerical axis
without an explicit design-to-information mapping. The assessment study can therefore
vary test length prospectively and ask whether Lee et al.'s conditional $I$-by-$J$ pattern
transfers. It cannot assume in advance that assessment lies above their crossovers.

## The contribution {#sec-ch21-contribution}

Our contribution is not the application of constrained Bayes and triple-goal estimation to
item response models. Those estimators are Ghosh's [-@ghosh_constrained_1992] and Shen and
Louis's [-@shen_triple-goal_1998], the three-goal framing is Shen and Louis's, and Lee et
al. [-@lee_improving_2025] have already shown, in multisite trials, that the value of a
flexible prior depends on how informative the data are — that semiparametric priors repay
their cost only when informativeness is high, and that below it a Gaussian prior paired
with the right posterior summary performs as well or better even when the true effect
distribution is not Gaussian.

What this work adds is a narrower, IRT-specific adaptation. It replaces the site-level
summary likelihood with the full binary-response likelihood, in which information is a
function of $\theta_p$ and of the item design. It must fix the latent location — and under
the 2PL the scale — before the recovered $G$ can be interpreted, and it must distinguish
the $I+1$ functionals identified by a finite Rasch test from the rest of a flexible
distribution [@sanmartin_bayesian_2011, Theorem 5]. Those requirements change the model,
the warrant for reading $G$, and the way reliability is manipulated.

The decision comparison also makes fixed-cut classification explicit as a fourth target.
That target is not new: Ghosh [-@ghosh_constrained_1992] motivates constrained Bayes with
above/below-cutoff and category decisions, and Paddock et al.
[-@paddock_flexible_2006] and Lockwood et al. [-@lockwood_uncertainty_2002] provide
adjacent threshold and classification precedents. The contribution is to include the
fixed substantive cut in this IRT comparison and keep it separate from ranking
(@sec-ch20), not to originate cut-score decision theory.

These structural differences motivate an empirical question rather than answer it.
Assessment reliability can be varied through test length, but whether the prior-by-summary
pattern found by Lee et al. changes under a full response likelihood is a realized result
assigned at @sec-ch21-not-claimed.

### What this statement deliberately does not claim {#sec-ch21-not-claimed}

Three restraints, each of which an earlier draft of this chapter violated.

It does not claim that the prior-and-summary decision *reverses* with reliability as a
finding of this work, or that assessments occupy a range beyond Lee et al.'s result. Lee
et al. show a prior-choice reversal on integrated squared-error loss conditional on $I$
and $J$, and they show it first. Whether that pattern transfers to a full IRT response
model as test length changes is a realized outcome, so the answer belongs to the companion
simulation study and not to this book (@sec-ch21-boundary).

It does not claim that adding items is cheaper than adding examinees. That is plausible
and it is not in the source; no argument here rests on it.

It does not claim that the trait variance is fixed by an identification constraint. Under
the Rasch model the metric is fixed by the model and $\sigma^2_\theta$ is a free parameter
(@sec-ch04); only the location needs constraining. Reliability is a design quantity
because test length moves the measurement error, not because the trait variance is pinned.

## Adjacent precedents {#sec-ch21-precedents}

Three works should be cited alongside rather than behind Lee et al., because each did part
of this before.

Paddock et al. [-@paddock_flexible_2006] combined flexible distributions with triple-goal
estimation in two-stage hierarchical models, and their results do not uniformly flatter
flexible methods (@sec-ch19). Lockwood et al. [-@lockwood_uncertainty_2002] supply the
Gaussian calibration relating rank recovery to a stability coefficient that @sec-ch20
uses; Lockwood et al. [-@lockwood_flexible_2018] later apply CB and TG coordinatewise in a
bivariate hierarchical model. Shen and Louis [-@shen_triple-goal_1998] are the origin of
the three goals and of the GR estimator, and are cited throughout Part VI.

## The boundary this chapter does not cross {#sec-ch21-boundary}

This chapter compares problems, models, and warrants. It does not report what any
simulation found — not this project's and not, beyond what is needed to state their
published conclusions, Lee et al.'s. Whether a flexible prior earns its cost at a given
reliability in an item-response setting is a realized outcome, and the companion
simulation study is the authority for it. The delineation above is complete without that
answer, which is the point: it establishes that the question is a different question, and
a different question is what the Associate Editor asked for.
