# Scope: Why Rasch, and What Extends {#sec-ch22}

```{r}
#| label: setup-ch22
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

The Associate Editor's second demand has two halves: justify restricting the work to a
unidimensional Rasch model, and say what happens outside it. Sufficiency supplies this
book's cleanest answer, but it is not the only logically possible design for studying
$G$.

**The chosen criterion is sufficiency.** A model that admits a sufficient statistic for
the person parameter admits conditional maximum likelihood, and CML estimates item
parameters without specifying $G$ (@sec-ch05). That gives this book a particularly clean
reference design: estimate the item side without a population-distribution model, then
vary $G$. Where person sufficiency fails, ordinary marginal calibration couples the item
and person sides through an assumed $G$. Alternatives such as fixed-effects joint
likelihood or externally known item parameters can still separate parts of the problem,
so sufficiency is a strong design reason for Rasch here, not a necessary condition for
every possible sensitivity study.

Reading the primary sources for this chapter moved the boundary. It does not fall where
"Rasch versus everything else" would put it.

One boundary this chapter once carried has moved inside the book, but the move does not
make the coverage symmetric. The theory spine remains Rasch-centred because its
cleanest arguments use sufficiency, CML, and the available semiparametric identification
result. Part VIII, **What Changes Under the 2PL**, is a bounded extension along three fault
lines: information and its allocation (@sec-ch24), reliability under uneven information
(@sec-ch25), and identification with a free scale (@sec-ch26). What remains here is
everything past the 2PL — polytomous response formats, multidimensionality, and diagnostic
classification.

## Why Rasch {#sec-ch22-why}

Three reasons, in decreasing order of how much this book leans on them.

**Sufficiency and CML.** The number-correct score is sufficient for $\theta_p$, so item
difficulties can be estimated by conditioning the person parameters out entirely. Masters
[-@masters_rasch_1982, p. 152] states the dichotomous case flatly: this is the *only* latent
trait model for dichotomously scored responses in which the number of successes is a
sufficient statistic for the person parameter. That uniqueness is the strongest reason
for the reference design used here. @sec-ch05 develops the estimator; what matters here
is that its existence gives this book a calibration point that does not specify $G$.

**No scale indeterminacy.** The logit metric fixes the unit, so only the location needs
constraining (@sec-ch04). This is what makes the semiparametric identification question
tractable at all: with one constraint, San Martín et al. [-@sanmartin_bayesian_2011] can say
exactly what a finite test identifies about $G$ (@sec-ch16). Under the 2PL a second
constraint is needed and the corresponding semiparametric result is not available
(@sec-ch22-2pl).

**The target context.** Curriculum-based measures and the short dichotomous instruments this
work is aimed at are where the 1PL is standard practice, and @sec-ch01 sets out their
measured reliabilities.

## What the Rasch-centred body already says about the 2PL {#sec-ch22-2pl}

The earlier parts introduce the 2PL at the places where it changes a Rasch argument; they
do not build a parallel theory from the ground up. This section consolidates those named
differences, and Part VIII follows the three that require a longer treatment.

Sufficiency is lost as soon as discriminations vary, because the score that would be
sufficient is weighted by the unknown $\lambda_i$ (@sec-ch03). A second constraint becomes
necessary, since location and scale are both free (@sec-ch04). Information changes shape:
doubling the discriminations does **not** quadruple the information, because the response
probabilities move too — the realized multiplier is 1.74 to 2.72 across $\theta$
(@sec-ch07). And one thing does not change at all: the shrinkage algebra of @sec-ch11 never
used sufficiency, so it holds verbatim.

What does not survive is the semiparametric identification result. San Martín
[-@sanmartin_identification_2016] marks the semiparametric 2PL with question marks in his
own summary table and says in the text that it is open; every identification statement this
book proves about $G$ is proved for the Rasch case. That is recorded as `C-010` because the
manuscript fits DPMs under the 2PL as well and does not disclose it.

## Polytomous models: the boundary is not where "Rasch versus the rest" puts it {#sec-ch22-polytomous}

The blueprint for this chapter expected the polytomous section to say that the shrinkage and
summary arguments carry over unchanged while information changes. That is true and it is not
the interesting part. Reading Masters and Muraki shows that **the sufficiency criterion cuts
straight through the polytomous family**, and it cuts in a place that matters.

**The partial credit model keeps everything.** Masters' model gives the probability of
scoring $x$ on an $m_i$-step item as an exponential of the accumulated step differences
[@masters_rasch_1982, eq. 10, p. 158]. Conditioning on the total count of completed steps
$r_n = \sum_i x_{ni}$ removes $\beta_n$ from the likelihood exactly as in the dichotomous
case [@masters_rasch_1982, eqs. 11–14, pp. 159–160], so $r_n$ is sufficient and CML is
available. Everything the CML part of the "why Rasch" argument buys is bought again here.
The step difficulties are estimated from the calibration responses without specifying
$G$; they are not free of the calibration sample.

**The graded response model loses it, and not because of the discrimination parameter.**
This is the sentence worth quoting to a reviewer. Masters [-@masters_rasch_1982, p. 155]
shows that the graded response form — built by differencing cumulative boundary
probabilities — "prevents the algebraic separation of person and item parameters" *even in
the absence of a discrimination parameter*, so no sufficient statistic exists for either
side. For this comparison the obstruction is the **category-boundary parameterization**,
not the slope. Samejima's graded response model parameterizes boundaries between regions
of a continuum; Masters parameterizes the difficulty of each step. Two different objects,
and only one of these two constructions separates. Masters' § 4 states the contrast, and a
direct check of Samejima [-@samejima_estimation_1969] confirms the cumulative
category-boundary construction. This local PCM-versus-GRM contrast does not establish a
universal boundary for every ordered-response parameterization.

**The generalized partial credit model loses it for the ordinary reason.** Muraki
[-@muraki_generalized_1992] adds a slope $a_j$ to the partial credit model, and says
directly that separability and minimal sufficient statistics are distinct properties of the
*Rasch* model, which is what permits CML [@muraki_generalized_1992, p. 160]. His derivative

$$\frac{\partial}{\partial\theta}P_{jk}(\theta) = a_j P_{jk}(\theta)
\left[k - \sum_{c} c\,P_{jc}(\theta)\right]$$

[@muraki_generalized_1992, eq. 13, p. 163] is the polytomous form of the same fact
@sec-ch07 records for the 2PL: the slope multiplies, but the category probabilities move
with it, so information does not scale as the slope squared.

One detail from Muraki bears directly on this book's programme and is easy to miss. His
estimation procedure is marginal maximum likelihood by EM, integrating $\theta$ against a
**normal** population density with Gauss–Hermite quadrature whose weights are approximately
standard normal ordinates [@muraki_generalized_1992, eqs. 17 and 22–23, pp. 165–167]. The
normality of $G$ is not a finding there; it is a population-model assumption in the
marginal likelihood. Gauss–Hermite quadrature is the computational method used to
approximate the corresponding integrals. The example shows how a conventional normal
model can be embedded in an implementation; it does not turn normality from a modelling
decision into a computational default.

## MIRT: the question gets harder, not easier {#sec-ch22-mirt}

Three constraints are needed rather than one — location, scale, and rotation — so the
identification argument of @sec-ch04 does not transfer, and the semiparametric question
@sec-ch16 answers for the Rasch case is not even posed in a settled form for the
multivariate one.

The shrinkage argument holds coordinatewise, and multivariate use is not absent. Shen and
Louis [-@shen_triple-goal_1998, p. 468] explicitly say that their approaches generalize
to multivariate unit-specific parameters and point to Ghosh for multivariate constrained
Bayes. Lockwood et al. [-@lockwood_flexible_2018, pp. 666–672] fit a bivariate
hierarchical model and apply CB and TG coordinatewise. These precedents close the
categorical “no multivariate form” claim.

What remains unsettled is narrower. A coordinatewise action depends on the chosen axes and
does not by itself define a joint ordering, a multivariate EDF target, or a
transformation-aware joint loss. This book does not construct a canonical joint CB/GR
action for vectors; that is the open problem it records.

Xu et al.
[-@xu_robustness_2025] test the three standard procedures for recovering item–trait
structure — exploratory item factor analysis with rotation, EM-based $L_1$ regularization,
and expectation model selection — under latent traits generated with varying skewness and
excess kurtosis, and report that non-normality generally lowers the $F_1$ score for
identifying item–trait relationships and raises the mean squared error of parameter
estimates, with method- and condition-specific exceptions, including cells in which EIFA
benefits. In the multidimensional case a misspecified $G$ can distort **which items load on
which dimension**, so the damage can reach the measurement model itself. Their discussion
also proposes more flexible latent distributions as a future response. This paper
therefore supports a robustness warning, not a literature-wide claim that the field
responds only with robustness assessment.

## Diagnostic classification: a bounded discrete analogue {#sec-ch22-cdm}

The blueprint expected this section to say that the flexible-$G$ programme is *replaced*
by a different question, because the latent variable is a discrete attribute profile
rather than a point on a continuum. That is too categorical. A distributional target has
a discrete analogue, but the estimands and identification objects are different.

In a restricted latent class model each respondent's profile $\boldsymbol\alpha$ is drawn
from a categorical distribution with proportions $\boldsymbol p = (p_{\boldsymbol\alpha})$,
and responses are conditionally independent given the profile [@gu_partial_2020, § 2]. Those
proportions are the discrete analogue of $G$: they are the latent distribution, and
recovering them is the analogue of the distributional goal.

Gu and Xu [-@gu_sufficient_2019, Theorem 1] give the sufficient **and necessary** condition
for identifying all DINA parameters, and it is a condition on the design alone: the
$Q$-matrix must be complete, each attribute must be required by at least three items, and
the columns of the submatrix below the identity block must be distinct
[@gu_sufficient_2019, Conditions 1–2, pp. 4–5]. When it holds, the maximum likelihood
estimates are consistent [@gu_sufficient_2019, Corollary 1].

The partial-identification result requires a distinction the first version of this chapter
missed. With **known item parameters**, Gu and Xu
[-@gu_partial_2020, Proposition 3.2] show that the sums of
$p_{\boldsymbol\alpha}$ over equivalence classes of profiles with identical
$\Gamma$-columns are identifiable. With **unknown item parameters**, inseparability alone
is not enough: their Definition 3.2 defines $\boldsymbol p$-partial identifiability,
Theorem 3.1 requires structural conditions C1 and C2, and Proposition 3.3 gives an
adjusted-$\Gamma$ route under the corresponding conditions. A raw inseparable
$\Gamma$-matrix therefore does not, by itself, identify the grouped proportions in the
full model.

The connection to @sec-ch16 is a limited methodological analogy. Gu and Xu's grouped
proportions are sums of masses over equivalence classes in a finite latent space. San
Martín et al.'s $I+1$ objects are integral functionals of a distribution on the real line,
not masses of a partition of that line. Both literatures warn against reporting more of a
latent distribution than a finite design identifies, but they do not identify
structurally identical objects and neither result is a template for the other.

What does not transfer is the estimators. Constrained Bayes rescales real-valued posterior
means to match a variance; triple-goal orders units and reads quantiles off an estimated
distribution function. Neither operation is defined on an unordered profile space, so
whether an analogue exists is open.

## The summary {#sec-ch22-summary}

```{r}
#| label: tbl-extensions
#| tbl-cap: "Does each part of the argument survive outside the Rasch model? Source: tables/T-extensions.rds."
#| echo: false
knitr::kable(tbl("T-extensions"))
```

Reading the column "Sufficient statistic for the person" down the table gives this book's
chosen answer to AE-2. The restriction is not conservatism about model complexity. It
selects models in which CML lets the cost of assuming $G$ be studied with the item side
calibrated without specifying $G$. That clean set contains the Rasch model and the partial
credit model, and it excludes the graded response model for a reason that, in the Masters
comparison, has nothing to do with discrimination parameters. Other study designs can
relax the necessity of this criterion, but they do not provide the same CML reference
point.

Two honest weaknesses. The 2PL is tracked at the points where the Rasch argument changes
even though it is outside that clean set; Part VIII makes those changes explicit, but the
semiparametric 2PL identification problem stays open and is disclosed as `C-010`. And
three cells in the table say "open in this book" for the polytomous families: what a finite
polytomous test identifies about $G$ is a question this work has not asked, and @sec-ch23
records it.
