# Open Problems {#sec-ch23}

```{r}
#| label: setup-ch23
#| include: false
tbl <- function(id) readRDS(file.path("..", "..", "tables", paste0(id, ".rds")))
```

This is the shortest chapter and it does the most to say what kind of document this is.
In the first edition it closed the Rasch-centred theory spine by listing what that argument
could not settle. In this edition it still closes that spine, but it does **not** end the
book. It is the hinge to Part VIII, **What Changes Under the 2PL**, and to Part IX, which
places the ledger beside realized evidence.

The hinge matters because the honest version of the argument in @sec-ch21 — that the
question changes when reliability is a design quantity — only carries weight if the
questions it *leaves* open are stated as plainly as the ones it closes.

Each entry says three things: what is actually known, what would settle it, and **whose
problem it is**. That last distinction matters. Some of these are places where this book had
the tools and did not do the work. Some belong to the companion simulation study by the
boundary this project set. And some are open in the field.

A note for the reader of this edition: the **entries** in the ledger below are preserved
in their first-edition state, because rewriting an open-problem record after the fact would
launder it. The surrounding navigation is updated. Readers asking how the item-model
change affects the theory should continue through @sec-ch24, @sec-ch25, and @sec-ch26.
Readers asking what became of the ledger should continue through @sec-ch27 and @sec-ch28
to @sec-ch29, which walks it item by item. Several entries are now settled by the companion
volumes, some in a sharper form than the entry asked for, and the evidence has added
entries of its own.

## Problems this book could have closed and did not {#sec-ch23-ours}

### The rank proposition outside its conditions {#sec-ch23-rank}

@prp-rank-reliability is stated with conditions because the general claim is false. What is
established is the leading-order statement — rank recovery depends on the model through the
stability coefficient — plus a measurable residual under a matched constant-error Gaussian
mapping and a Rasch computation on one grid of shapes.

Three things are unknown. Whether the residual's behaviour is general: on the displayed grid
the shape gap rises to a peak near $\bar w = 0.915$ and then declines, and the decline to
zero in the perfect-information limit is proved only for continuous working shapes. Whether
the ordering result of @sec-ch12 — that in a common-form Rasch model differential shrinkage
cannot reorder — survives the three regimes @sec-ch20 names as vulnerable: different test
forms, incomplete designs, and item uncertainty carried into the posterior. And whether the
matched-reliability construction is the right comparison at all, since it holds $\bar w$
fixed while the shapes differ, which is one of several defensible matchings.

The first two are finite computations. They were not run.

### How far the incompatibility theorem generalizes {#sec-ch23-incompat}

@thm-incompatible is an existence statement, and it is stated that way because an earlier
draft asserted a universal one and was refuted (@sec-ch17). The example is two independent
non-degenerate overlapping normals; the degenerate coincident case where the actions agree
is exhibited alongside it.

What is not known is the boundary. For which families, and under what separation of the
posterior means relative to their spreads, do the WSEL and ISEL actions coincide? The
question has a clean form — characterize the set of posterior ensembles on which the
midpoint-quantile action equals the vector of posterior means — and it is not answered here.
A useful partial answer would be the exchangeable equal-variance case, which is the one
applied work most often has.

### What a finite polytomous test identifies about $G$ {#sec-ch23-polytomous}

@sec-ch22 establishes that the partial credit model keeps sufficiency and the graded response
model does not. It does not ask the semiparametric question. San Martín et al.'s result —
that, under the anchored Rasch conditions, $I+1$ displayed integral evaluations of $G$
generate the observable law subject to their identities (@sec-ch16) — has no counterpart
here for a test of $I$ items with $m_i$ steps each.

The conjecture worth stating is that the count should depend on the total number of steps
rather than the number of items, since that is what the sufficient statistic counts. It is a
conjecture. Nothing in this book supports it beyond the analogy.

### Location conventions and interpretable functionals of $G$ {#sec-ch23-alpha}

The earlier version of this section asked whether the location constraint changes the
elicited prior on occupied-cluster count $K$. It does not. The implemented Rasch model
centres the item parameters and does not transform the DPM atoms; even if all atoms were
translated by one draw-specific constant, their cluster memberships and occupied count
would remain exactly unchanged.

A narrower question remains. Alternative location representatives, and the additional
scale transformation required under a 2PL model, change the numerical locations and
spreads used to describe $G$. This book does not characterize how priors on
decision-relevant functionals — such as tail mass beyond a fixed substantive cut,
variance, or separation between mixture components — transform across those conventions,
or which such summaries are invariant. The open object is therefore an interpretable
functional of $G$, not $K$.

### What the normal working model costs on short tests {#sec-ch23-working}

@sec-ch11 derives the closed-form posterior under a Gaussian working model and V8 verifies it
against numerical integration. The working model is an approximation to a posterior whose
error variance depends on $\theta$, and the approximation degrades as the test shortens. The
size of that error at the test lengths this work is aimed at — a dozen items or fewer — is
not quantified. It is a one-dimensional numerical exercise and it was not done.

## Problems the boundary assigns elsewhere {#sec-ch23-boundary}

Two questions in this book's neighbourhood belong to the companion simulation study, and
saying so here is not evasion but the same boundary @sec-ch21 draws.

**Whether Lee et al.'s prior-by-information pattern transfers to the IRT design.** Lee et
al. show a conditional pattern in $I$ and $J$. Chapter 21 does not place assessment on the
same numerical axis without a mapping. What happens under a full item-response likelihood
as test length changes is a realized outcome.

**Whether any of the theory's preferences survive Monte Carlo error.** Every statement in
this book is about what is true of a model, not about what a finite simulation can
distinguish.

## Problems open in the field {#sec-ch23-field}

### Finite-$N$ behaviour of the DPM posterior under weak identification {#sec-ch23-finiteN}

This is the deepest of them. Under the anchored Rasch conditions, a finite test's observable
law is generated by the $I+1$ displayed integral evaluations of $G$, subject to their
identities, while guarantees for flexible deconvolution estimators are asymptotic.
The literature read for this book does not give a finite-sample account of how posterior
uncertainty in unidentified directions affects decisions based on those identified
functionals.

@sec-ch22 supplies a bounded comparison, not a solution. In restricted latent class
models, grouped masses over finite equivalence classes can be identified when the relevant
known-item or unknown-item structural conditions hold. Those masses are not partitions of
continuous $G$, and San Martín's functionals already are a finite-design answer of a
different kind. The useful open question is which identified Rasch functionals are
decision-relevant and how a flexible posterior's nonidentified directions affect their
finite-sample interpretation.

### A multivariate constrained Bayes and triple-goal {#sec-ch23-mirt}

Multivariate use already exists. Shen and Louis state that their approaches generalize to
multivariate unit-specific parameters, and Lockwood et al. apply CB and TG coordinatewise
in a bivariate hierarchical model (@sec-ch22). What is not supplied by those precedents is
a canonical **joint** action: a transformation-aware loss, ordering or multivariate EDF
target that is not an arbitrary collection of coordinatewise decisions. The open problem
is to define and compare that joint target, not to invent multivariate CB/GR from nothing.

### Misspecification of the measurement model {#sec-ch23-measurement}

Every result in this book conditions on the item model being right. The whole programme asks
what it costs to assume the wrong $G$ while the Rasch or 2PL likelihood is correct.

Nothing here addresses the opposite failure, and there is reason to think it matters more.
Xu et al.'s finding in the multidimensional case (@sec-ch22) is that non-normality can distort
which items load on which dimension — the misspecified population distribution damaging the
measurement model, not just the recovered distribution. If that coupling has a unidimensional
analogue, then the separation this book relies on — item side estimable without $G$, so the
cost of $G$ is isolable — is a property of correctly specified item models and not of the
estimator.

That would not invalidate anything proved here. It would relocate it.

## A problem that is really a literature audit {#sec-ch23-audit}

@sec-ch13 shows that a bimodal $G$ can produce a unimodal sum-score distribution — total
variation $0.035$ on the worked case, with six spread items. The inference "the score
distribution looks normal, therefore the trait distribution is" is therefore invalid, and it
is common.

What is unknown is whether any *published* conclusion about latent non-normality rests on it.
Answering that means reading a body of applied work with a specific question, which is a
different kind of labour from the rest of this book and was not undertaken. It is listed
because it is the item on this list most likely to change what someone does tomorrow.

## The list {#sec-ch23-summary}

```{r}
#| label: tbl-open-problems
#| tbl-cap: "What this book does not settle. Source: tables/T-open-problems.rds."
#| echo: false
knitr::kable(tbl("T-open-problems"))
```

Five of the nine are marked as this book's own. That count records responsibility, not
difficulty. Some are bounded numerical checks; others require a new characterization or
semiparametric identification argument whose difficulty has not been established. A
reader deciding whether to trust @sec-ch21's contribution statement is entitled to know
which questions the book scoped out, without being told that they are all tractable.

Three of the remaining four are genuinely open in the field, and one of those three — the
finite-$N$ meaning of a DPM posterior for $G$ under weak identification — is the question
that would do most to justify or undercut the entire flexible-prior programme this book
describes.

## Where the ledger leads {#sec-ch23-forward}

This chapter ends the Rasch-centred theory spine, not the volume. The next route depends
on the reader's question:

- **What changes when discriminations are free?** Read @sec-ch24 for information,
  @sec-ch25 for reliability, and @sec-ch26 for identification and the DPM. Together they
  are Part VIII, **What Changes Under the 2PL**.
- **What did the programme's evidence settle?** Read @sec-ch27 for the simulation record,
  @sec-ch28 for the real-test record, and @sec-ch29 for the formal disposition of this
  ledger.
